---
title: "braves-rebuild"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Checking in on the Atlanta Braves Rebuild

=== How is the Atlanta Braves' rebuild going? ===

- Intro -- set up situations that led to Braves rebuild: On July 29th, 2014, the Atlanta Braves were a half game behind the Washington Nationals in the NL East. The Braves had won the division handily the year prior and reached the NL Wild Card game the year before that. But the team won just 18 of its next 49 games down the stretch in 2014. This included a miserable 4-18 stretch that saw them eliminated from the postseason and saw their GM at the time, Frank Wren, fired.
- Had signed multiple players to long-term deals, including Freddie Freeman to an eight-year, $135 million extension prior to 2014. Heyward and Simmons were also contract recipients around the same time.


- September 22nd, 2014, FRANK WREN FIRED. JOHN HART IS INTERIM GM; NAMED PREZ BASEBALL OPS A MONTH LATER.
- THEN, prior to 2015 season, the trades:
  -- Jason Heyward, Craig Kimbrel, Justin Upton, Melvin Upton, Evan Gattis
  -- Kris Medlen non-tendered.

2015 season start:
  -- Alex Wood traded mid-2015. 
  -- Only real recognizable names from previous years were Freddie Freeman & Julio Teheran. 

2015 offseason:
 -- Andrelton Simmons trade!
 -- Shelby Miller trade!

Now their only good players are Freddie Freeman and Ender Inciarte.


############
- The only question for the Atlanta Braves this year was whether they'd challenge the single-season record for losses. 
- The team has been in rebuild mode since the 2014-2015 offseason. That's when new GM John Coppolella and President of Baseball Operations John Hart traded away Jason Heyward, Craig Kimbrel, Justin Upton, Melvin Upton, and Evan Gattis. The following offseason he traded away Andrelton Simmons then shocked the baseball world by turning Shelby Miller (and, it must be noted, Gabe Speier) into Dansby Swanson, Ender Inciarte, and Aaron Blair.

Hart built the devastating Indians teams of the mid-'90's, .560 winning percentage during his tenure there, six AL Central titles, two WS appearances. From there he went to Texas where he was GM for one year before moving into an advisory role. He remained there through 2013 after seeing the Rangers rise from irrelevance to perennial contenders. 


- How do you judge a rebuild?? WELL HOW DO YOU !?!?!?!?
-- You can't go by wins in the standings. Those are not the priority for the current team.
-- You can't go by quality of the major-league roster.
-- And you can't judge by future wins. Since they, uh, aren't here now.
-- How well are they set up now, for the future?
-- How many future WAR do they have under their control, and for how long?
  -- WHAT'S THEIR LIKELY OPENING DAY ROSTER FOR 2017, AND HOW GOOD IS IT?!
  -- What is the state of their farm system?
  -- Braves had six players on Baseball America's Top 100 list in mid-July: 
    - SS Dansby Swanson #7, SS Ozzie Albies #17, 
  -- As recently as July 19th, Keith Law ranked their farm system #1: http://www.talkingchop.com/2016/7/19/12227724/keith-law-puts-ranks-braves-farm-system-1-in-his-midseason-system-rankings
  -- Of course the rankings may look different now, with the Yankees' huge haul at the trade deadline.
-- How well did they do in each of their transactions? (maybe .... ? doing this would be a shit load of work)
-- COULD JUST TALK ABOUT MOVES MADE DURING THE 2015 SEASON?!?!!?
--- Bud Norris, Jhoulys Chacin, Lucas Harrell, Kelly Johnson, Jason Grilli.


**********

- How Should Atlanta Braves Fans Judge the Rebuild?
- Examining the Rebuild in Atlanta


Atlanta Braves fans are suffering through another dismal season. The team is mired in last place with only a few bright spots on the major league roster. Many fans' hearts and minds stray to the Gwinnett and Mississippi Braves' rosters. The 2013 NL East division championship feels like a decade ago, never mind the 94-win season that preceded it.

The team's fans know the reasons behind their frustration. The organization is rebuilding, trying to put together another fourteen-year run. Right now the rosters in Gwinnett County, GA and Pearl, MS are more important than the one in Atlanta.

So fans understand. But eventually this strategy needs to pay off. There a lot of history at stake, not to mention pride and a brand-new ballpark. Two years into the regime of President of Baseball Operations John Hart and General Manager John Coppolella, how should fans feel about their progress?

==

What should these suffering fans pay attention to?

Time




The heads of the front office, John Hart as the President of Baseball Operations and John Coppolella as the General Manager, 



The Braves fired general manager Frank Wren towards the end of the 2014 season. Despite making the playoffs in 2010, 2012, and 2013, the team had just been eliminated from the postseason and would finish the year with a Pythagorean record of .481. The team installed John Hart and John Coppolella as President of Baseball Operations and General Manager, respectively. 

Apparently the time was right for a rebuild, as the 2014-2015 seasons saw a flurry of activity. They traded away stars Jason Heyward, Craig Kimbrel, Justin Upton, and Melvin Upton. They also got rid of breakout folk hero Evan Gattis. In the middle of the 2015 season, the team traded starter Alex Wood. After the season, they traded Andrelton Simmons and, most notably, Shelby Miller. Most of these deals brought back prospects or rookies.  

Believe it or not, but they didn't get rid of all their players. They still had to field 25 guys each night. While tinkering around the edges of the roster they kept the services of Julio Teheran and Freddie Freeman, signed Nick Markakis to a four-year deal, and traded for Nick Swisher and Michael Bourn.



This sell-off 



the team replaced him with John Hart as President of Baseball Operations and John Coppolella as the GM. They began 


===

How Should Atlanta Braves Fans Feel About the Rebuild? 

==Team Run Differential==

Of course fans want the team to win games. But run differential tells a team's story better than wins and losses do. Players foul a ball off their foot and are done for the year. Players hit home runs with the bases empty and the team down by four instead of with two men on and the game tied. Or the umpire calls a critical pitch a strike instead of a ball, ending the game instead of walking in the winning run.

The largest component of run differential is the players on the field. I don't know whom the Braves will field next year, but for right now, relying on Freddie Freeman is a safe bet. After somewhat of a down year in 2015, he'll finish this year around 6 WAR. Since 2011 he trails only Paul Goldschmidt and Joey Votto in WAR and matches about evenly with Anthony Rizzo. He's under contract through 2021.

Besides him, the team has Ender Inciarte who's wrapping up a 4-win season. Inciarte was everything fans thought he'd be, a league-average hitter who plays stellar defense at an up-the-middle position. He's also 25 years old and under team control through 2020.

These are position players you build a team around. On the other side of the ball, though things aren't so rosy. Julio Teheran is the team's nominal ace on the strength of a 2.9-win year, but he doesn't have the track record fans should feel excited about, and his 102 xFIP- suggests he's more of a league-average starter. This may be why GM John Coppolella told Buster Olney the team plans to focus on starting pitching this offseason [http://www.mlbtraderumors.com/2016/09/coppolella-on-swanson-free-agency-manager.html].

They may not have to look far. In Eric Longenhagen's words, the Braves farm system is "loaded with arms". Four of which appeared on Baseball America's mid-season Top 100: Sean Newcomb. Kolby Allard. Mike Soroka. Touki Toussaint. Add position players Dansby Swanson and Ozzie Albies and you start getting Braves fans excited.

Fans should also feel good about _how_ the team acquired their talent. After all, at FanGraphs we generally value process over results. We know about the eye-popping Shelby Miller trade. In addition to that incredible move, Longenhagen says "[The Braves] have done a tremendous job capitalizing on every asset available. The top of the farm system is loaded, but they've added relevant pieces by trading scrapheap pickups like Lucas Harrell. They maneuvered their draft pool quite well this year and got three monster prep arms in Ian Anderson, Joey Wentz, and Kyle Muller. That's a sign of an efficient organization."

If your mantra is process over results, you should be excited about the Braves position.

==Time==

People hate waiting for a team to be good, especially when it was so good so recently. Luckily we have a standard against which we can judge the Braves: three years.

The Chicago Cubs and Houston Astros provided us with this timeline. Both teams installed new regimes before the 2012 season. Both teams immediately tanked and remained below a .450 Pythagorean winning percentages for the next three years. Both teams' fortunes turned in 2015, resulting in playoff baseball. Both teams continue to contend this year.

```{r}
pythag_data <- read_csv("~/Downloads/braves rebuild - Sheet1 (2).csv")

fmt_labels <- function(x) sprintf("%.3f", x)

non_braves_rebuild <- pythag_data %>%
  filter(Team == "Cubs" |
         Team == "Astros")

ggplot(data=non_braves_rebuild, aes(x=Season, y=PythagPCT, color=Team)) + geom_line() + geom_point() + labs(x="Season", y="Pythagorean Win%") + theme(text=element_text(family = "Lato-Regular")) + scale_y_continuous(labels=fmt_labels) + ggtitle("Astros and Cubs Pythagorean Winning Percentage, 2007-2016")

```

In this respect the Braves are on track. Hart and Copollela fully took over after the 2014 season. Their Pythagorean record dropped immediately to under .400 and remained around that area this year:

```{r}
ggplot(data=pythag_data, aes(x=Season, y=PythagPCT, color=Team)) + geom_line() + geom_point() + labs(x="Season", y="Pythagorean Win%") + theme(text=element_text(family = "Lato-Regular")) + scale_y_continuous(labels=fmt_labels) + ggtitle("Astros, Braves, and Cubs Pythagorean Winning Percentage, 2007-2016")
```

Relative to the first rebuilding year, the Braves improved a bit this year. That's more encouraging than the Astros, who slid in their second rebuilding year. But it's not as good as the Cubs, who improved in their second rebuilding year by more than the Braves did.

What does 2017 hold for Atlanta? I don't know, but based on what happened in Chicago and Houston, fans should be okay with more losing. Contention would be a surprise based on good luck, akin to the Minnesota Twins' 2015 season, rather than a legitimate change, based on what the Twins front office _thought_ about their 2015 season.

Wins in Atlanta aren't guaranteed. Hart and Coppolella still have to build a team that can win. Chicago and Houston drafted some solid prospects (Kris Bryant and Kyle Schwarber in Chicago; Carlos Correa, Lance McCullers, and Alex Bregman in Houston) and hit on some solid trades (Jake Arrieta and Anthony Rizzo in Chicago, Ken Giles in Houston). Houston also benefitted from decisions made by the prior regime, most notably Jose Altuve and George Springer.

Braves fans have my permission to get grumpy if the team doesn't improve dramatically in 2018.

==Payroll Flexibility==

The third aspect of a rebuilding team is payroll flexibility. I define this as the gap between a team's spending limit and their committed payroll. Flexibility allows GMs to retain better players for longer while absorbing the impact of poor roster decisions. Flexibility also allows the team's executives to, should they choose to keep some of it as profit, drive a BMW Series 7 to work instead of a Nissan Altima.

We can't know a team's spending limit, but we can make an educated guess. Cot's Contracts has team Opening Day payroll information going back to 2000. We can find the team's highest payroll in terms of 2016 dollars and use that as the most they're willing to spend. From that, we subtract the team's Opening Day payroll in the third year of their rebuild. That gives us a good take on ownership's flexibility.

<div class="table-container" style="max-width: 600px; margin-top: 0px; margin-right: auto; margin-left: auto; background-color: white;"><div id="FG" style="position: absolute; height: 30px; width: 30px; margin-top: 3px; margin-left: 5px; background: url(&quot;http://www.fangraphs.com/blogs/wp-content/uploads/2015/08/FG_logo_transparency.png&quot;) 0% 0% / 100% no-repeat;"></div><div class="table-title" id="title-FG" style="min-height: 22px; text-align: center; vertical-align: middle; padding: 7px 50px; font-family: Lato, Arial; color: white; font-size: 18px; word-wrap: break-word; line-height: 1; background-color: rgb(82, 174, 38);">Payroll Flexibility of Recently Rebuilt Teams</div><table border="1" class="fg-table" style="font-family: Lato, Arial; max-width: 600px; width: 100%; border: 1px solid black; border-collapse: collapse; text-align: center; margin-bottom: 0px;"><tr class="table-header" style="text-align: center; color: white; background-color: rgb(80, 80, 80);"><td class="d3-th" column="0" style="border: 1px solid black;">Team</td><td class="d3-th" column="1" style="border: 1px solid black;">Year with Highest Payroll (Prior to Rebuild)</td><td class="d3-th" column="2">Highest Payroll Prior to Rebuild*</td><td class="d3-th" column="3">Third Year of Rebuild</td><td class="d3-th" column="4">Opening Day Payroll in Third Year of Rebuild*</td><td class="d3-th" column="5">Payroll Flexibility in Third Year of Rebuild</td></tr><tr class="table-rows" id="row-0" onmouseover="this.bgColor='#FDC60D'" onmouseout="this.bgColor=&quot;#FFFFFF&quot;" bgcolor="#FFFFFF" style="border: 1px solid black;"><td column="0" class="d3-td" style="padding-left: 5px; text-align: left;">Cubs</td><td column="1" class="d3-td">2010</td><td column="2" class="d3-td">$158,794,900.00</td><td column="3" class="d3-td">2014</td><td column="4" class="d3-td">$92,677,368.00</td><td column="5" class="d3-td">$66,117,532.00</td></tr><tr class="table-rows" id="row-1" onmouseover="this.bgColor='#FDC60D'" onmouseout="this.bgColor=&quot;#FFFFFF&quot;" bgcolor="#FFFFFF" style="border: 1px solid black;"><td column="0" class="d3-td" style="padding-left: 5px; text-align: left;">Astros</td><td column="1" class="d3-td">2009</td><td column="2" class="d3-td">$115,355,983.68</td><td column="3" class="d3-td">2014</td><td column="4" class="d3-td">$51,495,516.00</td><td column="5" class="d3-td">$63,860,467.68</td></tr><tr class="table-rows" id="row-2" onmouseover="this.bgColor='#FDC60D'" onmouseout="this.bgColor=&quot;#FFFFFF&quot;" bgcolor="#FFFFFF" style="border: 1px solid black; background-color: rgb(253, 198, 13);"><td column="0" class="d3-td" style="padding-left: 5px; text-align: left;">Braves</td><td column="1" class="d3-td">2003</td><td column="2" class="d3-td">$139,179,203.77</td><td column="3" class="d3-td">2017</td><td column="4" class="d3-td">$85,600,000.00**</td><td column="5" class="d3-td">$53,579,203.77</td></tr></table><div class="notes" style="padding: 3px 10px; font-size: 12px; font-family: Lato, Arial; border-left: 1px solid black; border-right: 1px solid black; border-bottom: 1px solid black;">* In 2016 Dollars. Source: Cot's Contracts<br>** Estimated. Source: Baseball-Reference</div></div>

Let's look at the Cubs. The third year of their rebuild was 2014. Prior to that year they'd recorded their highest payroll in 2010: $158.8 million. At the time, this was their theoretical spending limit. (Although they couldn't have expressed it in 2016 dollars, we do so here so we can compare to the upcoming Braves team.) The team opened the season with a payroll of $92.7 million. Therefore their flexibility that year was about $66 million.

The Braves fall a bit short here. They project to open the 2017 season with about $10 million less wiggle room than the Cubs and Astros did in the third year of their rebuild. That money needs to accommodate a few roster spots; A.J. Pierzynski, Jim Johnson, Eric O'Flaherty, Gorden Beckham, and Emilio Bonifacio are all free agents. Shedding Matt Kemp's salary via trade would serve the team well.

[[conclusion]]








For example, the New York Yankees' payroll flexibility next year is an incredible $113,217,384. Per Cot's Contracts their highest opening day payroll, adjusted to 2016 dollars, was in 2005 when they spent $256,217,384. But according to Baseball-Reference, the team has $143,000,000 in commitments next year, including options and projected arbitration salaries. $113 million can pay a lot of baseball players for one year. 

The Braves' payroll flexibility for 2017 is a respectable $53,579,203. Cot's tells us their highest Opening Day payroll occurred in 2003 at $139,179,203. Baseball-Reference estimates the team's 2017 payroll at $85,600,000. They could reduce that by 22% by trading Matt Kemp. 


 

===







before league-minimum players and arbitration salaries. The team has only four arbitration-eligible players: two second-timers (Arodys Vizcaino and Anthony Recker) and two first-timers (Ender Inciarte and Paco Rodrigiuez). Inciarte may get a large raise but the other players will not, especially considering Vizcaino and Rodriguez both missed this season with Tommy John Surgery.



Right now they have four arbitration-eligible players: Arodys Vizcaino (second time), Anthony Recker (second time), Ender Inciarte (first time), and Paco Rodriguez (first time). Work by Matt Perez at Camden Depot suggests the former two players can expect a 20% raise while the latter two can expect a 50% increase. Combined with these players' 2016 salaries, this work yields 2017 salaries of: 


Vizcaino: 897,500 + 20% = $1,077,000
Recker: 507,500 + 20% = $609,000
Inciarte: 523,000 + 50% = $784,500
Rodriguez: 520,500 + 50% = $780,750

= additional $3,251,350.

+ (other minimum-value players?)




====






that sequencing can affect team Thus far I've talked only about runs scored and runs allowed. I haven't mentione wins or losses. This was intentional. Yes, wins are the entire point of baseball. But so much of winning games is unpredictable sequencing: your fireballing closer hangs a curveball with a one-run lead, your star first baseman strikes out with the bases loaded, or the opposing shortstop taps a sinking liner that tips off the glove of your center fielder.

Presidents of Baseball Ops and General Managers can't predict these outcomes. What they can do, what fans should hold them accountable for, is build a team that scores as many runs as possible while allowing as few as possible, to minimize the impact poor sequencing can have. If a closer hangs a curveball, hopefully it's with a three-run lead. If a first baseman strikes out with the bases loaded, hopefully his team is up by three runs instead of tied. If the opposing shortstop taps a sinking liner, hopefully none of his teammates are on base.

Time.  






If the team doesn't take a big leap forward in 2018, fans will have a right to start grumbling. I'm not saying they'll contend. The Washington Nationals are still set up well, the Phillies have some exciting young talent, and the Mets have Tim Tebow. But the Braves should score more runs, and allow fewer runs, than in any of the previous three years.

-- Team run differential. 

--- how to actually judge this? 2016? projections for 2017? projected openiny day roster + top prospects?     - SS Dansby Swanson #7, SS Ozzie Albies #17, LHP 


-- Salary flexibility. Owners and general managers would prefer to build teams as cheaply as possible. The former, so they can drive to work in a BMW 7 Series instead of a Nissan Altima. The latter, so they can take on salary when necessary to improve the team's chances while absorbing a few mistakes along the way.

To figure out how much salary flexibility the Braves have, we can look at the highest Opening Day payroll they've ever had, translate it into 2016 dollars, and find the gap between that number and their committed 2017 payroll. The latter might change as the team makes moves this offseason. And it might change even more with the new Collective Bargaining Agreement due this winter. But it's a good starting point.

Cot's Contracts tells us the Braves' highest Opening Day payroll occurred in 2014 at $112,008,731. The Bureau of Labor Statistics reports buying power has decreased by 2% since then, leading to a 2016 equivalent of $109,768,556. That's the most the Braves have spent on a team. For reference their Opening Day Payroll this year was $86,580,791, giving them $23,187,765 of wiggle room.

For 2017 the team has only $50,826,000 committed before league-minimum players and arbitration salaries. Arbitration-eligible players include Arodys Vizcaino (second time), Anthony Recker (second time), Ender Inciarte (first time), and Paco Rodriguez (first time). Work by Matt Perez at Camden Depot suggests the former two players can expect a 20% raise while the latter two can expect a 50% increase. Combined with these players' 2016 salaries, this work yields 2017 salaries of: 


Vizcaino: 897,500 + 20% = $1,077,000
Recker: 507,500 + 20% = $609,000
Inciarte: 523,000 + 50% = $784,500
Rodriguez: 520,500 + 50% = $780,750

= additional $3,251,350.

+ (other minimum-value players?)







Run differential breaks down into a few 
-- Roster Depth

-- Compare to the Astros and the Cubs, two best-case scenarios.
--- Cubs lost shitloads of games for three years under Epstein. (2012-2014) before becoming good.

Year,Cubs Pythag PCT
2010,.450
2011,.432
-- Hoyer & Epstein Hired
2012,.401
2013,.438
2014.438
2015,.556
2016,.660 (so far)


--- Astros lost shitloads of games for three years under Luhnow (2012-2014) before becoming good.

Year - PythPCT
2010 - .419
2011 - .382
-- Luhnow Hired
2012 - .364
2013 - .351
2014 - .438
2015 - .574
2016 - .493 (so far)



--- Braves so far:
2010 - .574
2011 - .524
2012 - .567
2013 - .604
2014 - .481
-- Hart & Copollela put in place
2015 - .376
2016 - .400 (so far)

The Braves stand out here in that their pre-regime change run was very successful.



So the going rate for rebuilds these days seems to be three years of ineptitude followed by contention. Since Hart / Copollela took control prior to the 2015 season, 2016 would be considered year two of their rebuild. And therefore, under the Cubs/Astros template, fans should expect the team to be good in 2018.

Will this happen!?!?!? What will the team look like then!?!?!?

**********




The Cubs dominated all of baseball this year. The Astros would have dominated the AL West but for plenty of ill-timed (for them) clutch hitting by the Texas Rangers.

New Stadium ... announced in Nov. 2013. Opening next year.


  -
  -
  -
  -

############

-- Judge the moves at the time they were made.




The Atlanta Braves shocked the baseball world last December 9th when they announced they'd traded Shelby Miller and Gabe Spier to Arizona for Dansby Swanson, Ender Inciarte, and Aaron Blair. The trade followed the exodus of big names Andrelton Simmons earlier that year and Justin Upton and Jason Heyward the year prior. 





-Heading into the season, it was an open question which team would be worse: The Phillies or the Braves.

Well the results are nearly in and the answer is the Braves, by a long shot. The team currently sits XX games behind Philly for the honor. They are scoring 3.92 Runs/Game, second-worst in the NL, while allowing 4.98 runs per game, third-worst in the NL.

The rebuild began in earnest 

-Shelby Miller trade

https://www.google.com/trends/explore?q=shelby%20miller
