
Headlines:

-Cubs Dominate MLB Surplus Value
-


... Intro ... what emotion do I want the reader to feel?

(Awe)

You're familiar with the concept of surplus value: a player's value to the team minus the salary they make. Value most often takes the form of WAR, but players can provide value in other ways -- attracting more fans to the stadium, selling jerseys, and so on.

It turns out the recently crowned Chicago Cubs not only have a core of young, cost-controlled talent beyond any team in baseball. They're also paying very little for that talent.

METHODOLOGY 

Surplus value under control has three parts: the number of years a player is under control, their WAR during that time period, and their salary during that time period. Below I describe the process I used to calculate each and any assumptions I made.

To determine the number of years a player is under control, I used team payroll information from Baseball Reference. To simplify my task, I assumed all contract options are exercised and no opt-outs are. I limited the time horizon to five years because I have WAR projections only for the next five years. For example, although Miguel Cabrera is under control for at least seven more years, my analysis only shows him as under control for the next five.

To project WAR I used a method Tom Tango proposed that weights past performance and applies an aging curve. The details are here: http://tangotiger.com/index.php/site/comments/war-marcels-...-warcels; read up if you're interested. The method isn't perfect, particularly when it comes to pitchers, but I trust Tango's methodology enough to go forward. For missing years I gave the player 0 WAR.

Projecting salary was trickier:

- For free-agent contract years, I information when available from Baseball Reference. I noticed they don't defer salary; for example, Chris Davis will make $17M next year, but $23M is listed on the Orioles' payroll page. I used the latter number.
- For 2017 arbitration salaries, I used projections from MLB Trade Rumors.
- For arbitrations years beyond 2017, I applied the 25%/40%/60% research described by Kevin Creagh at The Point of Pittsburgh (http://www.thepointofpittsburgh.com/calculating-mlb-arbitration-percentages/). For example, Jonathan Schoop is projected to earn $3.4M in 2017, his Arb-1 year. To project his Arb-2 salary, I calculated his free-agent AAV as (3.4 / .25), then multiplied by .4 to get a projection of $5.44M. I then used $5.44M to project his Arb-3 salary at $8.16M. 
- For Arb-4 years, I used 70% of free-agent salary. For simplicity's sake, I did not identify Arb-4 players and then apply the 20%/33%/50%/70% progression that Creagh found. The progression in my analysis goes 25%/40%/60%/70%.
- For pre-arb salaries, I used the 2017 league minimum of $535k. I did not step up each year as the CBA dictates.

The toughest part was projecting arbitration salaries for players who will be pre-arb in 2017. No arbitration projection yet exists for these fellas. I used $2.2M, which is the average Arb-1 salary given out in 2016, according to MLBTR's tracker (http://www.mlbtraderumors.com/arbtracker2016). The 25/40/60/70 rubric kicked in for subsequent arbitration projections.

Using averages in situations like these has its perils. For example, Kris Bryant will get far more than $2.2M in his first arbitration season. So in the following analysis, his surplus value is drastically overstated. But short of projecting Arb-1 for every player uniquely, this was the best I could do. I'm open to other methods if you'd like to share them in the comments.

With years of control, WAR under control, and salary under control, I calculated a player's surplus value as (8.5 * WAR) - Salary. I inflated the value of WAR by 5% each year and inflated salary by 3% each year.

===SECTION HEADING===

OK, now we can get to the results, which are simply this: the Cubs dominate the entire sport of Major League Baseball in terms of talent they control and the surplus value they're getting from that talent. It's not even close.

Without talking about salary at all, you can see Chicago controls the most amount of talent by far:

```{r}
ggplot(data=team_war, aes(years_of_control, war_under_control_per_year, label=Team)) + geom_point() + geom_text(aes(label=Team),hjust=0, vjust=0) + labs(x="Avg. years of control  per player (5 years max)", y="Avg. WAR under control per year") + ggtitle("MLB Team Control of Talent, 2017-2021") + xlim(2.5,5)
```

When it comes to talent under control, the Cubs lead the sport and no one else is close. The Mets come the closest to the Cubs in terms of overall talent under control, but they average half a year less of control per player. Their contention window is much smaller. Nationals fans should be happy; your team controls a similar amount of talent but for longer, on average, than your division rivals. Red Sox, Dodgers, Astros, and Giants fans also have reason to be excited. 

But don't get it twisted. The Cubs dominate everyone else and they likely will for years to come.

When we add salary into the equation we can calculate the surplus value each team has under control. Here again the Cubs lead the other teams by a large margin:

```{r}
cubs_tbl <- team_war %>%
  select(-num_players, -war_under_control, -years_of_control, -salary_under_control, - war_under_control_per_year,
         -war_per_player_per_year, -surplus_value_per_year, -mm_per_war_per_year) %>%
  arrange(desc(surplus_value_under_control))

cubs_tbl

```

[[explain how the $2.2M makes this gap look much larger than it probably really is]]


This table inflates the value of many pre-arb players, particularly good ones like the Cubs have


, because Here's where the Arb-1 projections of Kris Bryant and many of his teammates come into play. I used a value of $2.2M as I explained above but Bryant, Addison Russell , Kyle Hendricks, and others on their team will make much more than this in their first year of arbitration.

have yet to hit arbitration. This makes them cheap regardless of how they perform on the field, but these players have sky-high WAR projections for the next five years.  


```{r}
cubs_position <- 
```

But not all surplus value is created equally. Some teams control theirs for a few years; others, for a long time. The following chart shows how the Cubs' average surplus value per year, along with the length of time they control it, compares to other teams:

```{r}

team_surplus_value <- warcels %>%
  group_by(Team) %>%
  summarize(avg_surplus_value_per_year = sum(surplus_value_under_control) / sum(years_of_control),
            avg_years_of_control = round(mean(years_of_control), 2))

ggplot(team_surplus_value, aes(avg_years_of_control, avg_surplus_value_per_year, Team)) + geom_point() + geom_text(aes(label=Team),hjust=0, vjust=0) + labs(x="avg. years of control", y="avg. surplus value per year (Millions USD)") + ggtitle("MLB Team Surplue Value Under Control, 2017 - 2021") + xlim(2.5,5)

```


Eight Cubs rank in the 90th percentile of 1,053-player. The most any other club has in this bracket is six -- shared by the Reds and Indians. The Red Sox, Astros, Dodgers, Twins, Phillies, Giants, and Blue Jays have five players each.

```{r}

teams_in_top_90 <- warcels %>%
  filter(surplus_value_under_control_rank >= 90) %>%
  group_by(Team) %>% 
  summarize(num_players_in_top_10_perc = n()) %>%
  arrange(desc(num_players_in_top_10_perc))

teams_in_top_90

```


These eight Cubs are:

```{r}

cubs_surplus_value <- warcels %>%
  filter(Team == "CHC", 
         surplus_value_under_control_rank >= 90.0) %>%
  mutate(salary_under_control = round(salary_under_control, 1)) %>%
  select(-Position, -Team, -war_per_year_of_control, -dollars_per_win_under_control)

cubs_surplus_value
```

Kris Bryant leads not only the Cubs, but all of MLB in controlled surplus value. His infield mates Anthony Rizzo and Addison Russell also rank highly. Now -- this list wildly underestimates Bryant's arbitration salaries. I declined to meaningfully project first-year arbitration salaries, which means his arbitration years come in at a very low $5M total. He'll certainly earn more than that. But that doesn't change the fact that he's 25 and projected for nearly 6 WAR per year for the next five years.

=== Cubs' Service Time Games with Bryant ... ? ===





