---
title: "Hof-voting-bayes"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(stringr)
#library(plotly)
library(RMySQL)
library(broom)
library(gamlss)
library(glue)

fgt = theme(
  panel.background = element_rect(color = "lightgrey", fill = "white"),
  axis.title = element_text(family = "Lato"),
  axis.text = element_text(family = "Lato"),
  legend.title = element_text(family = "Lato"),
  legend.text = element_text(family = "Lato"),
  legend.background = element_rect(fill = "white"),
  legend.key = element_rect(fill = "white"),
  plot.title = element_text(family = "Lato"),
  panel.grid.major = element_line(size = .1, color = "lightgrey"),
  panel.grid.minor = element_line(size = 0),
  #strip.background = element_rect(fill = "#50ae26"),
  strip.text = element_text(family="Lato")
  )

fg_db <- dplyr::src_mysql(dbname = Sys.getenv("FG_DB"), 
                            host = Sys.getenv("FG_HOST"), 
                            user = Sys.getenv("FG_USER"),
                            password = Sys.getenv("FG_PW"))

get_voting_results <- function(year) {
  # get player data and (if available) HOF voting results for a particular year
  # also get total ballots cast that year
  url <- str_c("https://www.baseball-reference.com/awards/hof_", year, ".shtml")

html <- read_html(url)

ballot_text <-
  html %>%
  html_nodes("li") %>% 
  html_text() %>%
  as_tibble() %>% 
  filter(str_detect(value, "total ballots")) %>%
  distinct(value) # needed because some years, like 2011, have a vote total listed twice

total_votes <- as.numeric(str_extract(ballot_text$value, "[:digit:]{3}"))

html %>%
    html_nodes("table") %>%
    .[[1]] %>%
    html_table(header = FALSE) %>%
    as_tibble() %>%
    mutate(vote_year = as.numeric(year),
           total_ballots = total_votes)
}

fix_name <- function(name) {
  # scrub player's name of weird characters that BB-Ref puts in their HOF voting data tables
  str_replace_all(str_replace(name, "y ", ""),
                  c(
                    "X-" = "",
                    "\\sHOF" = ""))
}

```


```{r}
# foo <-
#   read_html("https://www.baseball-reference.com/awards/hof_2011.shtml") %>% 
#   html_nodes("li") %>% 
#   html_text() %>%
#   as_tibble() %>% 
#   filter(str_detect(value, "total ballots"))
# 
# as.numeric(str_extract(foo$value, "[:digit:]{3}"))[1]
# 

```


get voting results so we can estimate beta prior parameters

for each player, get: 
- year on ballot
- playing position
- JAWS
- # votes that year
- # total votes 


```{r}
hof_voting_data <- seq(1990, 2018) %>%
  map_df(get_voting_results)
```

clean and transform data so it can be processed correctly
```{r}
hof_voting_data_clean <-
  hof_voting_data %>% 
  dplyr::select(Name = X2,
                num_votes_received = X4,
                year_on_ballot = X3,
                vote_year,
                total_ballots,
                jaws = X11,
                jpos = X12
  ) %>%
  filter(!Name %in% c("Name", "")) %>% # remove informational and spacer rows
  mutate(Name = fix_name(Name),
         num_votes_received = as.numeric(num_votes_received),
         year_on_ballot = as.numeric(str_extract(year_on_ballot, "[:digit:]{1,2}")),
         jaws = as.numeric(jaws),
         jpos = as.numeric(jpos),
         jaws_plus = round(100 * (jaws / jpos), 0)) %>%
  dplyr::select(-jaws, -jpos)

# get ballot and stat information for those on the next ballot. 
year_to_predict <- 2019
current_url <- str_c("https://www.baseball-reference.com/awards/hof_", year_to_predict, ".shtml")
current_hof_voting <-
  read_html(current_url) %>%
  html_nodes("table") %>%
  .[[1]] %>%
    html_table(header = FALSE) %>%
    as_tibble() %>%
    mutate(vote_year = as.numeric(year_to_predict)) %>%
  dplyr::select(Name = X2,
         year_on_ballot = X3,
         vote_year,
         jaws = X10,
         jpos = X11) %>%
  filter(!Name %in% c("Name", "")) %>% # remove informational and spacer rows
  mutate(Name = fix_name(Name),
         year_on_ballot = as.numeric(str_extract(year_on_ballot, "[:digit:]{1,2}")),
         jaws = as.numeric(jaws),
         jpos = as.numeric(jpos),
         jaws_plus = round(100 * (jaws / jpos), 0)) %>%
  dplyr::select(-jaws, -jpos)

# get to-date voting information from HOF Tracker and combine with BB-Ref stats, and join to previously scraped data
current_voting_data <-
  read_csv("2019 BBHOF Tracker/Summary-Table 1.csv")
known_votes <- as.numeric(current_voting_data[2,3]) + as.numeric(current_voting_data[1,3])
voting_info_to_date <-
  current_voting_data %>%
  filter(row_number() >= 5) %>% # remove metadata
  dplyr::select(Name = X1,
                vote_share = X2) %>%
  mutate(vote_share = as.numeric(str_extract(vote_share, "[:digit:]{1,3}.{1}")) / 100,
         total_ballots = known_votes,
         num_votes_received = ceiling(vote_share * total_ballots)) %>%
  inner_join(current_hof_voting, by="Name") %>%
  dplyr::select(Name, num_votes_received, year_on_ballot, vote_year, total_ballots, jaws_plus)
master <- bind_rows(hof_voting_data_clean, voting_info_to_date)

```

```{r}

# TODO: use JAWS relative to avg JAWS @ position, not JAWS itself.
# TODO: now add in information from 2019. can i do this? their voting_years will be 1. they will prob. need better priors .. 


# %>%
#   group_by(Name) %>%
#   summarize(total_votes = sum(num_votes_received),
#             total_ballots = sum(total_ballots),
#             years_on_ballot = n())

fit <- gamlss(cbind(num_votes_received, total_ballots - num_votes_received) ~ 0 + log(year_on_ballot) * log(jaws_plus),
              data = master,
              family = BB(mu.link = "identity"))

# td <- tidy(fit)
# 
# mu_0 <- td$estimate[1]
# mu_PA <- td$estimate[2]
# sigma <- exp(td$estimate[3])

# plot prior distirbutions for players' HOF voting, based on their year on the ballot and JAWS. 
# note these are just the medians.
# TODO: might be better to quantize JAWS and show ACTUAL CURVES for players with 10, 50, and 100 JAWS to get the point across.

# sigma <- fitted(fit, "sigma")[1]
# crossing(jaws = c(25, 50, 75, 100),
#          year_on_ballot = c(1, 5, 10)) %>%
#   augment(fit, newdata = .) %>%
#   rename(mu = .fitted) %>%
#   crossing(x = seq(0, 1, .1)) %>%
#   mutate(alpha = mu / sigma,
#          beta = (1 - mu) / sigma,
#          density = dbeta(x, alpha, beta)) %>%
#   ggplot(aes(x, density, color = factor(year_on_ballot), group = jaws)) +
#   geom_line() +
#   labs(x = "Estimated HOF",
#        y = "Prior density",
#        color = "AB",
#        lty = "Batting hand")


# plot medians of prior distributions of vote share totals, given player's jaws & the year they are on the ballot 
# TODO: I want to show the actual distro curves but haven't figured out how to do that successfully. some code for it is in my NOTES
master %>%
    augment(fit, newdata = .) %>%
    rename(mu = .fitted) %>%
    mutate(sigma = fitted(fit, "sigma")[1],
           alpha0 = mu / sigma,
           beta0 = (1 - mu) / sigma,
           conf_low = qbeta(.025, alpha0, beta0),
           conf_high = qbeta(.975, alpha0, beta0)) %>%
    ggplot(aes(year_on_ballot, mu, color = jaws_plus, group = jaws_plus)) +
    geom_line() +
    #geom_ribbon(aes(ymin = conf_low, ymax = conf_high), linetype = 2, alpha = .1) +
    labs(x = "Year on Ballot",
         y = "Estimate Prior HOF Vote Share (Median)",
         color = "JAWS+") +
  geom_hline(yintercept = 0.75, linetype="dashed", color="red") +
  ylim(0, 1) + fgt



```


```{r}
# td <- tidy(fit)
# 
# mu <- fitted(fit, parameter = "mu")
# sigma <- fitted(fit, parameter="sigma")

# combine prior distributions with likelihoods to produce posterior distributions.
# simplify visualizations to most likely values + credible intervals instead of probability distributions
est_hof_voting <- 
  master %>%
  ungroup() %>%
  mutate(mu = predict(fit, what = "mu", newdata = master),
         sigma = predict(fit, what = "sigma", newdata = master, type = "response"),
         alpha0 = mu / sigma,
         beta0 = (1 - mu) / sigma,
         alpha1 = alpha0 + num_votes_received,
         beta1 = beta0 + total_ballots - num_votes_received,
         est_hof_vote_share = alpha1 / (alpha1 + beta1),
         low_credible = qbeta(.025, alpha1, beta1), 
         high_credible = qbeta(.975, alpha1, beta1)) %>%
  mutate(name_year = paste(Name, vote_year, sep = " ")) %>%
  arrange(desc(est_hof_vote_share))

# ken griffey jr. prior vs. posterior
ggplot() + geom_histogram(binwidth = 0.009, aes(y=..density..)) +
  stat_function(fun = dbeta,
                args=list(shape1 = 0.2426830, shape2 = 1.932842),
                aes(color = "Prior")) +
  stat_function(fun = dbeta,
                args=list(shape1 = 437.24268, shape2 = 4.932842),
                aes(color="Posterior")) +
  xlim(0, 1) + labs(x="HOF Vote Share") + 
  ggtitle("Ken Griffey Jr. HOF Voting:  Prior Expectations vs. Posterior Probabilities") +
  fgt

#mariano rivera: prior vs. posterior



#compare two players' posterior distributions visually
ggplot() + geom_histogram(binwidth = 0.009, aes(y=..density..)) +
  stat_function(fun = dbeta,
                args=list(shape1 = 337.4489, shape2 = 86.77561),
                aes(color = "Trevor Hoffman")) +
  stat_function(fun = dbeta,
                args=list(shape1 = 47.4489, shape2 = 376.77561),
                aes(color="Billy Wagner")) +
  xlim(0, 1) + labs(x="HOF Vote Share") + 
  ggtitle("HOF Vote Share Derby: Two Great Closers") +
  fgt

```

```{r, fig.height=3}
ggplot(est_hof_voting %>% filter(vote_year == 2019), aes(est_hof_vote_share, reorder(Name, est_hof_vote_share))) +
  geom_point() +
  geom_errorbarh(aes(xmin = low_credible, xmax = high_credible)) +
  labs(y="", x="Estimated HOF Vote % with 95% Credible Interval", 
       caption = "Sources: \nPast voting: Baseball Reference\n2019 voting: bbhoftracker.com") +
  ggtitle("Projected HOF Voting, Class of 2019",
          subtitle = glue("After {known_votes} votes tallied")) + fgt +
  geom_vline(xintercept = 0.75, linetype = "dashed", color = "green") +
  geom_vline(xintercept = .05, linetype = "dashed", color = "red")
```




ahhhh but i can't estimate a prior for players w/o any voting results ... can i??

they DO have some ACTUAL voting results (likelihood). perhaps I can use a naive prior? beta with alpha = 1 and beta = 1? or comparable players?



do empirical analysis for first-year players. they haven't had any results to observe yet ... or have they?

