---
title: "Hof-voting-bayes"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(stringr)
# library(plotly)
library(RMySQL)
library(RMariaDB)
library(rvest)
library(broom)
library(gamlss)
library(glue)

fgt <- theme(
  panel.background = element_rect(color = "lightgrey", fill = "white"),
  axis.title = element_text(family = "Lato"),
  axis.text = element_text(family = "Lato"),
  legend.title = element_text(family = "Lato"),
  legend.text = element_text(family = "Lato"),
  legend.background = element_rect(fill = "white"),
  legend.key = element_rect(fill = "white"),
  plot.title = element_text(family = "Lato"),
  panel.grid.major = element_line(size = .1, color = "lightgrey"),
  panel.grid.minor = element_line(size = 0),
  # strip.background = element_rect(fill = "#50ae26"),
  strip.text = element_text(family = "Lato")
)

fg_db <-
  DBI::dbConnect(RMariaDB::MariaDB(),
    dbname = Sys.getenv("FG_DB"),
    host = Sys.getenv("FG_HOST"),
    user = Sys.getenv("FG_USER"),
    password = Sys.getenv("FG_PW"),
    port = 5001,
    ssl.ca = "~/Documents/skysql_fangraphs-db.pem"
  )

get_voting_results_bayes <- function(year) {
  # get player data and (if available) HOF voting results for a particular year
  # also get total ballots cast that year
  url <- str_c("https://www.baseball-reference.com/awards/hof_", year, ".shtml")

  html <- read_html(url)

  ballot_text <-
    html %>%
    html_nodes("li") %>%
    html_text() %>%
    as_tibble() %>%
    filter(str_detect(value, "total ballots")) %>%
    distinct(value) # needed because some years, like 2011, have a vote total listed twice. also years like 1967 had 2 elections

  total_votes <- as.numeric(str_extract(ballot_text$value, "[:digit:]{3}"))
  # print(glue("year: {year}, total ballots: {total_votes}"))

  html %>%
    html_nodes("table") %>%
    .[[1]] %>%
    html_table(header = FALSE) %>%
    as_tibble() %>%
    mutate(
      vote_year = as.numeric(year),
      total_ballots = total_votes
    )
}

fix_name <- function(name) {
  # scrub player's name of weird characters that BB-Ref puts in their HOF voting data tables
  str_replace_all(
    name,
    c(
      "X-" = "",
      "\\sHOF" = ""
    )
  )
}

plot_prior_posterior <- function(Name, alpha0, beta0, alpha1, beta1, est_hof_vote_share) {
  ggplot() +
    geom_histogram(binwidth = 0.1, aes(y = ..density..)) +
    stat_function(
      fun = dbeta,
      args = list(shape1 = alpha0, shape2 = beta0),
      aes(color = "Prior")
    ) +
    stat_function(
      fun = dbeta,
      args = list(shape1 = alpha1, shape2 = beta1),
      aes(color = "Posterior")
    ) +
    xlim(0, 1) +
    labs(
      x = "Estimated HOF Vote Share",
      color = "Distribution",
      caption = "Sources:\nPast voting: Baseball Reference\n2019 voting: bbhoftracker.com"
    ) +
    ggtitle(glue("{Name} HOF Voting 2022"),
      subtitle = glue("Estimated vote share: {est_hof_vote_share * 100}%")
    ) +
    fgt +
    geom_vline(xintercept = 0.75, linetype = "dashed", color = "green", lwd = 1.1) +
    geom_vline(xintercept = .05, linetype = "dashed", color = "red", lwd = 1.1)
}
```


get voting results so we can estimate beta prior parameters

```{r}
hof_voting_data_bayes <- seq(1980, 2021) %>%
  map_df(get_voting_results_bayes)
```

clean and transform data so it can be processed correctly
```{r}
hof_voting_data_bayes_clean <-
  hof_voting_data_bayes %>%
  dplyr::select(
    Name = X2,
    num_votes_received = X4,
    year_on_ballot = X3,
    vote_year,
    total_ballots,
    bj_hof_monitor = X6,
    best_seven_war = X10,
    position = X39,
    G_pitch = X31,
    GS = X32,
    SV = X33
  ) %>%
  filter(!Name %in% c("Name", "")) %>% # remove informational and spacer rows
  mutate(
    Name = fix_name(Name),
    num_votes_received = as.numeric(num_votes_received),
    year_on_ballot = as.numeric(str_extract(year_on_ballot, "[:digit:]{1,2}")),
    bj_hof_monitor = as.numeric(bj_hof_monitor),
    G_pitch = as.numeric(G_pitch),
    GS = as.numeric(GS),
    SV = as.numeric(SV),
    best_seven_war = as.numeric(best_seven_war),
    position = str_extract(position, "[:digit:]"),
    position_class = case_when(
      position == 1 & GS / G_pitch >= 0.5 ~ "Starting pitcher",
      position == 1 & GS / G_pitch <= 0.5 ~ "Relief pitcher",
      TRUE ~ "Position player"
    )
  ) %>%
  dplyr::select(-position)

# get to-date voting information from BB HOF Tracker and combine with BB-Ref stats, and join to previously scraped data
current_voting_data <-
  read_csv("~/Documents/2022 BBHOF Tracker/Summary-Table 1.csv")
known_votes <- as.numeric(current_voting_data[2, 3]) + as.numeric(current_voting_data[1, 3])

ballot_2022 <-
  read_html("https://www.baseball-reference.com/awards/hof_2022.shtml") %>%
  html_table(header = TRUE) %>%
  pluck(3) %>%
  dplyr::select(
    Name = 2, year_on_ballot = 3, vote_share = 4,
    bj_hof_monitor = 5, bj_hof_standards = 6, career_length = 7, career_rWAR = 8,
    best_war_years = 9, jaws = 10, jpos = 11, G_bat = 12, AB = 13, R = 14, H = 15, HR = 16, RBI = 17,
    SB = 18, BB = 19, BA = 20, OBP = 21, SLG = 22, OPS = 23, OPS_plus = 24,
    W = 25, L = 26, ERA = 27, ERA_plus = 28, WHIP = 29, G_pitch = 30, GS = 31, SV = 32, IP = 33,
    H_allowed = 34, HR_allowed = 35, BB_allowed = 36, SO = 37, position = 38
  ) %>%
  mutate(
    vote_year = 2022,
    Name = fix_name(Name),
    year_on_ballot = as.numeric(str_extract(year_on_ballot, "[:digit:]{1,2}")),
    vote_share = as.double(str_extract(vote_share, "[:digit:]{1,3}.[:digit:]")),
    bj_hof_monitor = as.numeric(bj_hof_monitor),
    bj_hof_standards = as.numeric(bj_hof_standards),
    career_length = as.numeric(career_length),
    career_rWAR = as.double(career_rWAR),
    best_war_years = as.double(best_war_years),
    jaws = as.numeric(jaws),
    jpos = as.numeric(jpos),
    jaws_plus = 100 * (jaws / jpos),
    G_bat = as.numeric(G_bat),
    AB = as.numeric(AB),
    R = as.numeric(R),
    H = as.numeric(H),
    HR = as.numeric(HR),
    RBI = as.numeric(RBI),
    SB = as.numeric(SB),
    BB = as.numeric(BB),
    BA = as.double(BA),
    OBP = as.double(OBP),
    SLG = as.double(SLG),
    OPS = as.double(OPS),
    OPS_plus = as.numeric(OPS_plus),
    W = as.numeric(W),
    L = as.numeric(L),
    ERA = as.double(ERA),
    ERA_plus = as.numeric(ERA_plus),
    WHIP = as.double(WHIP),
    G_pitch = as.numeric(G_pitch),
    GS = as.numeric(GS),
    SV = as.numeric(SV),
    IP = as.double(IP),
    H_allowed = as.numeric(H_allowed),
    HR_allowed = as.numeric(HR_allowed),
    BB_allowed = as.numeric(BB_allowed),
    SO = as.numeric(SO),
    position = factor(str_extract(position, "[:digit:]")), # encode position as where they played the most often
  ) %>%
  filter(!is.na(year_on_ballot)) %>%
  ungroup() %>%
  mutate(position_class = case_when(
    position == 1 & GS / G_pitch >= 0.5 ~ "Starting pitcher",
    position == 1 & GS / G_pitch <= 0.5 ~ "Relief pitcher",
    TRUE ~ "Position player"
  ))



# attach current voting into to player's statistics and combine with data scraped above
# the variable 'hof_voting_data_future_clean' is contained in hof-voting.Rmd; i used it for the machine learning piece
voting_info_to_date <-
  current_voting_data %>%
  filter(row_number() >= 5) %>% # remove metadata
  dplyr::select(
    Name = `...1`,
    vote_share = `...3`
  ) %>%
  mutate(
    vote_share = as.numeric(str_extract(vote_share, "[:digit:]{1,3}.{1}")) / 100,
    total_ballots = known_votes,
    num_votes_received = ceiling(vote_share * total_ballots)
  ) %>%
  inner_join(ballot_2022, by = "Name") %>%
  dplyr::select(Name, num_votes_received, year_on_ballot, vote_year, total_ballots, bj_hof_monitor,
    best_seven_war = best_war_years, G_pitch, GS, SV, position_class
  )

master_bayes <-
  bind_rows(hof_voting_data_bayes_clean, voting_info_to_date) %>%
  group_by(Name) %>%
  arrange(year_on_ballot) %>%
  mutate(prev_year_vote_share = lag(num_votes_received) / lag(total_ballots)) %>%
  dplyr::select(
    Name, num_votes_received, year_on_ballot, vote_year, total_ballots, bj_hof_monitor,
    SV, position_class, prev_year_vote_share
  ) %>%
  ungroup()
```

for players returning to the ballot, voting heavily depends on the prior year's voting (r^2 = 0.93)
i can prove this with a scatterplot
use this info to fit model for returning players and estimate their 2019 vote share distribution 

```{r}

# remove saves column; it's not needed to fit the model, and gamlss() will complain about it having NA values
repeat_players <-
  master_bayes %>%
  filter(
    year_on_ballot != 1,
    !is.na(prev_year_vote_share)
  ) %>%
  dplyr::select(-SV) %>%
  ungroup()

fit <- gamlss(cbind(num_votes_received, total_ballots - num_votes_received) ~ prev_year_vote_share,
  data = repeat_players,
  family = BB(mu.link = "logit")
)

# td <- tidy(fit)
mu <- fitted(fit, parameter = "mu")
sigma <- fitted(fit, parameter = "sigma")

est_hof_repeat_players <-
  repeat_players %>%
  mutate(
    mu = mu,
    alpha0 = mu / sigma,
    beta0 = (1 - mu) / sigma,
    est_prior_hof_vote_share = alpha0 / (alpha0 + beta0),
    alpha1 = alpha0 + num_votes_received,
    beta1 = beta0 + total_ballots - num_votes_received,
    est_hof_vote_share = round(alpha1 / (alpha1 + beta1), 3),
    low_credible = qbeta(.025, alpha1, beta1),
    high_credible = qbeta(.975, alpha1, beta1),
    min_probable = qbeta(0.001, alpha1, beta1),
    max_probable = qbeta(.999, alpha1, beta1)
  ) %>%
  filter(vote_year == 2021) # TODO: two vote years with 2021 for some reason ...

# plot priors and posteriors for interesting players
est_hof_repeat_players %>%
  dplyr::select(Name, alpha0, beta0, alpha1, beta1, est_hof_vote_share) %>%
  filter(Name %in% c("Curt Schilling", "Barry Bonds", "Todd Helton", "Omar Vizquel")) %>%
  pmap(plot_prior_posterior)
```


all of the above works only for players returning to the ballot
for first-year players, I fit separate models for starting pitchers, position player, and relievers
each class uses somewhat different stats

```{r echo=FALSE, message=FALSE, warning=FALSE}

master_starters <-
  master_bayes %>%
  filter(
    year_on_ballot == 1,
    position_class %in% c("Starting pitcher", "Position player")
  ) %>%
  dplyr::select(-SV, -prev_year_vote_share)

fit_starters <-
  gamlss(cbind(num_votes_received, total_ballots - num_votes_received) ~ bj_hof_monitor,
    data = master_starters,
    family = BB(mu.link = "logit")
  )

# td_starters <- tidy(fit_starters)
mu_starters <- fitted(fit_starters, parameter = "mu")
sigma_starters <- fitted(fit_starters, parameter = "sigma")

est_hof_first_year_starters <-
  master_starters %>%
  mutate(
    mu = mu_starters,
    alpha0 = mu / sigma_starters,
    beta0 = (1 - mu) / sigma_starters,
    alpha1 = alpha0 + num_votes_received,
    beta1 = beta0 + total_ballots - num_votes_received,
    est_hof_vote_share = round(alpha1 / (alpha1 + beta1), 3),
    low_credible = qbeta(.025, alpha1, beta1),
    high_credible = qbeta(.975, alpha1, beta1)
  ) %>%
  filter(vote_year == 2021)

est_hof_first_year_starters %>%
  dplyr::select(Name, alpha0, beta0, alpha1, beta1, est_hof_vote_share) %>%
  filter(Name %in% c("Mark Buehrle", "Tim Hudson")) %>%
  pmap(plot_prior_posterior)
```


```{r}

master_relievers <-
  master_bayes %>%
  filter(
    year_on_ballot == 1,
    position_class == "Relief pitcher",
    vote_year == 2021
  ) %>%
  dplyr::select(-prev_year_vote_share)

fit_relievers <-
  gamlss(cbind(num_votes_received, total_ballots - num_votes_received) ~ bj_hof_monitor + SV,
    data = master_relievers,
    family = BB(mu.link = "logit")
  )

# td_relievers <- tidy(fit_relievers)
mu_relievers <- fitted(fit_relievers, parameter = "mu")
sigma_relievers <- fitted(fit_relievers, parameter = "sigma")

est_hof_first_year_relievers <-
  master_relievers %>%
  mutate(
    mu = mu_relievers,
    alpha0 = mu / sigma_relievers,
    beta0 = (1 - mu) / sigma_relievers,
    alpha1 = alpha0 + num_votes_received,
    beta1 = beta0 + total_ballots - num_votes_received,
    est_hof_vote_share = round(alpha1 / (alpha1 + beta1), 3),
    low_credible = qbeta(.025, alpha1, beta1),
    high_credible = qbeta(.975, alpha1, beta1)
  ) %>%
  filter(vote_year == 2022)

est_hof_first_year_relievers %>%
  dplyr::select(Name, alpha0, beta0, alpha1, beta1, est_hof_vote_share) %>%
  filter(Name %in% c("Mariano Rivera", "Darren Oliver")) %>%
  pmap(plot_prior_posterior)
```


```{r, fig.height=3}

# plot posterior curves of some players, to show what pre-flattening looks like

ggplot() +
  geom_histogram(binwidth = 0.1, aes(y = ..density..)) +
  stat_function(
    fun = dbeta,
    args = list(shape1 = 48.59796, shape2 = 209.2028),
    aes(color = "Scott Rolen")
  ) +
  stat_function(
    fun = dbeta,
    args = list(shape1 = 181.7613, shape2 = 76.03944),
    aes(color = "Roger Clemens")
  ) +
  stat_function(
    fun = dbeta,
    args = list(shape1 = 91.63613, shape2 = 166.1647),
    aes(color = "Fred McGriff")
  ) +
  stat_function(
    fun = dbeta,
    args = list(shape1 = 21.97254, shape2 = 235.8282),
    aes(color = "Andruw Jones")
  ) +
  stat_function(
    fun = dbeta,
    args = list(shape1 = 155.6397, shape2 = 102.1611),
    aes(color = "Larry Walker")
  ) +
  xlim(0, 1) +
  labs(
    x = "Estimated HOF Vote Share",
    color = "Player",
    caption = "Sources:\nPast voting: Baseball Reference\n2019 voting: bbhoftracker.com"
  ) +
  ggtitle(glue("HOF Voting 2019: Example Posterior Distributions",
    subtitle = glue("After {known_votes} ballots revealed")
  )) +
  fgt +
  geom_vline(xintercept = 0.75, linetype = "dashed", color = "green", lwd = 1.1) +
  geom_vline(xintercept = .05, linetype = "dashed", color = "red", lwd = 1.1)
```


```{r}
# final plot of all voting
est_hof_all <-
  bind_rows(est_hof_repeat_players, est_hof_first_year_starters, est_hof_first_year_relievers)

ggplot(est_hof_all, aes(est_hof_vote_share, reorder(Name, est_hof_vote_share))) +
  geom_point() +
  geom_errorbarh(aes(xmin = low_credible, xmax = high_credible)) +
  labs(
    y = "", x = "Estimated HOF Vote Share with 95% Credible Interval",
    caption = "Sources:\nPast voting: Baseball Reference\n2022 voting: bbhoftracker.com"
  ) +
  ggtitle("Projected 2022 HOF Voting",
    subtitle = glue("As of {Sys.Date()}")
  ) +
  fgt +
  geom_vline(xintercept = 0.75, linetype = "dashed", color = "green", lwd = 1.1) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray", lwd = 1.1) +
  geom_vline(xintercept = .05, linetype = "dashed", color = "red", lwd = 1.1)

# condense to just a table of point estimates, chances of induction, and chances of elimination
est_hof_all %>%
  mutate(
    `Estimated Vote Share (%)` = round(est_hof_vote_share * 100, 3),
    `Chance of Induction (%)` = round(1 - pbeta(.75, alpha1, beta1), 3) * 100,
    `Chance of Elimination (%)` = round(pbeta(.049, alpha1, beta1), 3) * 100
  ) %>%
  dplyr::select(Name, contains("%")) %>%
  arrange(desc(`Estimated Vote Share (%)`))
```



```{r}

# evaluate model using actual 2019 voting

hof_voting_2019 <- seq(2019, 2019) %>%
  map_df(get_voting_results_bayes)

hof_voting_2019_clean <-
  hof_voting_2019 %>%
  dplyr::select(
    Name = X2,
    actual_vote_share = X5
  ) %>%
  filter(!Name %in% c("Name", "")) %>% # remove informational and spacer rows
  mutate(
    Name = fix_name(Name),
    actual_vote_share = as.numeric(str_extract(actual_vote_share, "[:digit:]{1,3}.[:digit:]")) / 100
  )

final_data <-
  est_hof_all %>%
  inner_join(hof_voting_2019_clean, by = "Name") %>%
  dplyr::select(Name, est_hof_vote_share, actual_vote_share) %>%
  mutate(residual = actual_vote_share - est_hof_vote_share)

bayes_model_r_sq <-
  summary(lm(actual_vote_share ~ est_hof_vote_share,
    data = final_data
  ))$r.squared %>%
  round(3)

bayes_model_rmse <-
  RMSE(final_data$est_hof_vote_share, final_data$actual_vote_share) %>%
  round(3)

ggplot(final_data, aes(est_hof_vote_share, actual_vote_share)) +
  geom_point() +
  geom_smooth(method = "lm") +
  fgt +
  geom_abline(linetype = "dashed") +
  labs(
    x = "Predicted Vote Share", y = "Actual Vote Share",
    caption = "Source: https://tht.fangraphs.com/2019-hall-of-fame-voting-edgar-martinez-roy-halladay-mariano-rivera/"
  ) +
  ggtitle("Evaluating Bayesian HOF Model",
    subtitle = glue("R^2: {bayes_model_r_sq}, RMSE = {bayes_model_rmse}")
  )

ggplot(final_data, aes(est_hof_vote_share, residual)) +
  geom_point() +
  fgt +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Predicted Vote Share", y = "Residual",
    caption = "Source: https://tht.fangraphs.com/2019-hall-of-fame-voting-edgar-martinez-roy-halladay-mariano-rivera/"
  ) +
  ggtitle("Evaluating Bayesian HOF Model: Residuals")
```
