---
title: "Hof-voting-bayes"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(stringr)
#library(plotly)
library(RMySQL)
library(broom)
library(gamlss)

fgt = theme(
  panel.background = element_rect(color = "lightgrey", fill = "white"),
  axis.title = element_text(family = "Lato"),
  axis.text = element_text(family = "Lato"),
  legend.title = element_text(family = "Lato"),
  legend.text = element_text(family = "Lato"),
  legend.background = element_rect(fill = "white"),
  legend.key = element_rect(fill = "white"),
  plot.title = element_text(family = "Lato"),
  panel.grid.major = element_line(size = .1, color = "lightgrey"),
  panel.grid.minor = element_line(size = 0),
  #strip.background = element_rect(fill = "#50ae26"),
  strip.text = element_text(family="Lato")
  )

fg_db <- dplyr::src_mysql(dbname = Sys.getenv("FG_DB"), 
                            host = Sys.getenv("FG_HOST"), 
                            user = Sys.getenv("FG_USER"),
                            password = Sys.getenv("FG_PW"))

get_voting_results <- function(year) {
  # get player data and (if available) HOF voting results for a particular year
  # also get total ballots cast that year
  url <- str_c("https://www.baseball-reference.com/awards/hof_", year, ".shtml")

html <- read_html(url)

ballot_text <-
  html %>%
  html_nodes("li") %>% 
  html_text() %>%
  as_tibble() %>% 
  filter(str_detect(value, "total ballots")) %>%
  distinct(value) # needed because some years, like 2011, have a vote total listed twice

total_votes <- as.numeric(str_extract(ballot_text$value, "[:digit:]{3}"))

html %>%
    html_nodes("table") %>%
    .[[1]] %>%
    html_table(header = FALSE) %>%
    as_tibble() %>%
    mutate(vote_year = as.numeric(year),
           total_ballots = total_votes)
}


```


```{r}
# foo <-
#   read_html("https://www.baseball-reference.com/awards/hof_2011.shtml") %>% 
#   html_nodes("li") %>% 
#   html_text() %>%
#   as_tibble() %>% 
#   filter(str_detect(value, "total ballots"))
# 
# as.numeric(str_extract(foo$value, "[:digit:]{3}"))[1]
# 

```


get voting results so we can estimate beta prior parameters

for each player, get: 
- year on ballot
- playing position
- JAWS
- # votes that year
- # total votes 


```{r}
hof_voting_data <- seq(first_training_year, last_training_year) %>%
  map_df(get_voting_results)
```

clean and transform data so it can be processed correctly
```{r}
hof_voting_data_clean <-
  hof_voting_data %>% 
  dplyr::select(Name = X2,
         num_votes_received = X4,
         year_on_ballot = X3,
         vote_year,
         total_ballots,
         jaws = X11
         ) %>%
  filter(!Name %in% c("Name", "")) %>% # remove informational and spacer rows
  mutate(
     Name = str_replace(Name, "y ", ""),
    Name = str_replace_all(Name, c(
      "X-" = "",
      "\\sHOF" = ""
    )),
    num_votes_received = as.numeric(num_votes_received),
         year_on_ballot = as.numeric(str_extract(year_on_ballot, "[:digit:]{1,2}")),
         jaws = as.numeric(jaws))


# TODO: now add in information from 2019. can i do this? their voting_years will be 1. they will prob. need better priors .. 
# TODO: can probably estimate their priors using JAWS as well as year on ballot.


# %>%
#   group_by(Name) %>%
#   summarize(total_votes = sum(num_votes_received),
#             total_ballots = sum(total_ballots),
#             years_on_ballot = n())

fit <- gamlss(cbind(num_votes_received, total_ballots - num_votes_received) ~ log(year_on_ballot),
              data = hof_voting_data_clean,
              family = BB(mu.link = "identity"))

td <- tidy(fit)

mu <- fitted(fit, parameter = "mu")
sigma <- fitted(fit, parameter="sigma")

# combine prior distributions with likelihoods to produce posterior distributions.
# simplify visualizations to most likely values + credible intervals instead of probability distributions
est_hof_voting <- hof_voting_data_clean %>%
  ungroup() %>%
  mutate(mu = mu,
         alpha0 = mu / sigma,
         beta0 = (1 - mu) / sigma,
         alpha1 = alpha0 + num_votes_received,
         beta1 = beta0 + total_ballots - num_votes_received,
         est_hof_vote_share = alpha1 / (alpha1 + beta1),
         low_credible = qbeta(.025, alpha1, beta1), 
         high_credible = qbeta(.975, alpha1, beta1)) %>%
  mutate(name_year = paste(Name, vote_year, sep = " ")) %>%
  arrange(desc(est_hof_vote_share))

ggplot(est_hof_voting %>% head(25), aes(est_hof_vote_share, reorder(name_year, est_hof_vote_share))) +
  geom_point() +
  geom_errorbarh(aes(xmin = low_credible, xmax = high_credible)) +
  labs(y="", x="Estimated HOF Vote %  with 95% Credible Interval") +
  ggtitle("THIS GRAPH NEEDS A FUCKING TITLE") + fgt

```




ahhhh but i can't estimate a prior for players w/o any voting results ... can i??

they DO have some ACTUAL voting results (likelihood). perhaps I can use a naive prior? beta with alpha = 1 and beta = 1? or comparable players?



do empirical analysis for first-year players. they haven't had any results to observe yet ... or have they?

