---
title: "manager-analysis"
output: html_document
---



```{r}
# library(tidyverse)
library(dplyr)
library(tidyr)
library(broom)
library(stringr)
library(RMySQL)
library(lubridate)
library(plotly)
library(glue)
library(styler)

fgt <- theme(
  panel.background = element_rect(color = "lightgrey", fill = "white"),
  axis.title = element_text(family = "Lato"),
  axis.text = element_text(family = "Lato"),
  legend.title = element_text(family = "Lato"),
  legend.text = element_text(family = "Lato"),
  legend.background = element_rect(fill = "white"),
  legend.key = element_rect(fill = "white"),
  plot.title = element_text(family = "Lato"),
  panel.grid.major = element_line(size = .1, color = "lightgrey"),
  panel.grid.minor = element_line(size = 0),
  # strip.background = element_rect(fill = "#50ae26"),
  strip.text = element_text(family = "Lato")
)

fg_db <- dplyr::src_mysql(
  dbname = Sys.getenv("FG_DB"),
  host = Sys.getenv("FG_HOST"),
  user = Sys.getenv("FG_USER"),
  password = Sys.getenv("FG_PW")
)

TWILIO_SID <- Sys.getenv("TWILIO_SID")
TWILIO_TOKEN <- Sys.getenv("TWILIO_TOKEN")
twilio_number <- Sys.getenv("twilio_number")

send_message <- function(msg) {
  tw_send_message(from = twilio_number, to = "5123505107", body = msg)
}
```

```{r}
get_team_managers <- function(team) {
  # this function takes a team as an input and returns a list of manager names, given the season is 2007 or later
  # and the team had only one manager that year

  url <- glue("https://www.baseball-reference.com/teams/{team}/index.shtml")

  tryCatch(
    url %>%
      read_html() %>%
      html_nodes("table") %>%
      .[[1]] %>%
      html_table(header = TRUE) %>%
      as_tibble() %>%
      dplyr::select(Season = Year, Manager = Managers) %>%
      filter(
        !str_detect(Manager, "and"),
        Season >= 2007
      ) %>%
      mutate(
        Manager = str_extract(Manager, "[A-Z].[:graph:]*"),
        Team = team
      ) %>%
      dplyr::select(Season, Team, Manager),
    error = function(e) {
      print(glue("No such URL for team abbreviation: {team}"))
    }
  )
}

# get managers for teams since 2007 that've had only one manager per year. replace a few team names to match FanGraphs'
team_managers <- c(
  "ARI", "ATL", "BAL", "BOS", "CHC",
  "CHW", "CIN", "CLE", "COL", "DET",
  "HOU", "KCR", "ANA", "LAD", "FLA",
  "MIL", "MIN", "NYM", "NYY", "OAK",
  "PHI", "PIT", "SDP", "SEA", "SFG",
  "STL", "TBD", "TEX", "TOR", "WSN"
) %>% map_df(get_team_managers) %>%
  mutate(Team = case_when(
    "ANA" ~ "LAA",
    "TBD" ~ "TBR"
  ))
```


- quantify good bullpen decisions by correlating k-bb% for the first 3 months of the season with average gmLI for the second half of the season
- player assumptions: player faced at least 25 batters in the first half. k-bb%+ is computed relative to others on team that year
- team assumptions: at least 5 relievers appeared for them 5 times in the 2nd half

get data

```{r}

players <- 
  fg_db %>%
  tbl("player_info") %>%
  filter(BISPosition == "RP") %>%
  dplyr::select(playerid = PlayerId, FirstName, LastName) %>%
  collect(n = Inf) %>%
  unite(player_name, FirstName, LastName, sep = " ")

teams <-
  fg_db %>%
  tbl("season_team") %>%
  filter(Season >= 2007) %>%
  dplyr::select(Season, TeamId, AbbName, W, L) %>%
  collect(n = Inf) %>%
  mutate(pct = W / (W + L))

reliever_projections <-
  fg_db %>%
  tbl("proj_pitching") %>%
  filter(Type == "Steamer",
         PlayerId %in% !!players$playerid) %>%
  collect(n = Inf) %>%
  mutate(k_perc = SO / TBF, 
         bb_perc = BB / TBF,
         proj_k_minus_bb = (k_perc - bb_perc) * 100) %>%
  dplyr::select(PlayerId, Season, k_perc, bb_perc, proj_k_minus_bb)


# get k-bb% for pitchers who faced at least 25 batters in the first half, computed relative to the other pitchers that year (who also
# faced at least 25 batters in the first half).
# reliever_perf2 <-
#   fg_db %>%
#   tbl("wpa_daily_pitching") %>%
#   filter(GameDate >= "2007-01-01") %>%
#   dplyr::select(GameDate, PlayerId, TeamId, SO, BB, TBF) %>%
#   collect(n = Inf) %>%
#   mutate(
#     Season = year(GameDate),
#     month = month(GameDate)
#   ) %>%
#   filter(month %in% c(3, 4, 5, 6)) %>%
#   group_by(Season, PlayerId) %>%
#   filter(sum(TBF) >= 25) %>%
#   summarize(
#     k_perc = sum(SO) / sum(TBF),
#     bb_perc = sum(BB) / sum(TBF)
#   ) %>%
#   mutate(first_half_K_BB_perc = round((k_perc - bb_perc) * 100, 1)) %>%
#   group_by(Season) %>%
#   mutate(first_half_K_BB_perc_plus = round(first_half_K_BB_perc / mean(first_half_K_BB_perc) * 100, 0))

# get plate appearances that began with a pitcher who was subbed in for another pitcher
wpa_data <- fg_db %>%
  tbl("wpa_log") %>%
  filter(
    !is.na(PitcherSubId),
    GameDate >= "2007-01-01"
  ) %>%
  collect(n = Inf)
```

K-BB%+ for 1st Half vs. gmLI for second half

```{r}

# get the average LI when entering a game, for pitchers in the 2nd half of the season July thru September
li_individual <-
  wpa_data %>%
  inner_join(players, by = c("P" = "playerid")) %>%
  mutate(
    Season = year(GameDate),
    game_month = month(GameDate)
  ) %>%
  filter(game_month %in% c(7, 8, 9)) %>%
  inner_join(teams, by = c("Season",
    "HomeTeamId" = "TeamId"
  )) %>%
  mutate(home_team = AbbName) %>%
  inner_join(teams, by = c("Season",
    "AwayTeamId" = "TeamId"
  )) %>%
  mutate(
    away_team = AbbName.y,
    inning_half = if_else(InningHalf == 0, "Top", "Bottom"),
    pitching_team = if_else(inning_half == "Top", home_team, away_team),
    pitching_team_score = if_else(inning_half == "Top", HomeScore, AwayScore),
    batting_team_score = if_else(inning_half == "Bottom", HomeScore, AwayScore),
    pitching_team_lead = pitching_team_score - batting_team_score
  ) %>%
  dplyr::select(Season, GameDate, game_month, home_team, away_team, inning_half,
    inning = Inning, outs = Outs, pitching_team_score, batting_team_score, pitching_team_lead, pitching_team, pitcher_playerid = P,
    pitcher_subbed_in = player_name, LI
  )

# find the mean gmLI for each pitcher in the second half and add in their first-half K-BB% data, who their manager was, and the team's performance that year
li_grouped <-
  li_individual %>%
  group_by(Season, pitching_team, pitcher_playerid, pitcher_subbed_in) %>%
  summarize(
    second_half_num_appearances = n(),
    second_half_mean_gmLI = round(mean(LI), 2)
  ) %>%
  inner_join(reliever_perf2, by = c("Season", "pitcher_playerid" = "PlayerId")) %>% # attach first-half k-bb rates to second-half gmLI stats
  inner_join(team_managers, by=c("Season", "pitching_team" = "Team")) %>%
  inner_join(teams, by=c("Season", "pitching_team" = "AbbName")) %>%
  dplyr::select(-W, -L, -TeamId)


# correlations should be done by YEAR and MANAGER, not team necessarily. 

manager_correlations <-
  li_grouped %>%
  ungroup() %>%
  split(.$Manager) %>%
  map(~ lm(second_half_mean_gmLI ~ first_half_K_BB_perc_plus, data = .)) %>%
  map(summary) %>%
  map_dbl("r.squared") %>%
  tidy() %>%
  mutate(r_squared = round(x, 2),
         r = sqrt(r_squared)) %>%
  dplyr::select(names, r_squared, r) %>%
  arrange(desc(r_squared))

manager_correlations




# for each season-team, identify its correlation between first-half K-BB% and second-half gmLI.
# this speaks to managers' rigidity
# correlations <-
#   li_grouped %>%
#   unite(season_team_manager, Season, pitching_team, Manager, sep = "-") %>%
#   split(.$season_team) %>%
#   map(~ lm(second_half_mean_gmLI ~ first_half_K_BB_perc_plus, data = .)) %>%
#   map(summary) %>%
#   map_dbl("r.squared") %>%
#   tidy() %>%
#   mutate(r_squared = round(x, 2)) %>%
#   dplyr::select(names, r_squared) %>%
#   arrange(desc(r_squared))

```

plot buck showalter's usage for each year he was in charge of the Orioles

```{r}
showalter_plt <-
  li_grouped %>%
  filter(Manager == "B.Showalter",
         pitching_team == "BAL") %>%
  ggplot(aes(first_half_K_BB_perc_plus, second_half_mean_gmLI, text = pitcher_subbed_in)) + geom_point() +
  labs(x="Reliever First-Half K-BB%+", y="Reliever Second Half Mean gmLI") +
  facet_wrap(vars(Season)) + ggtitle("Decision-Making in the Second Half, Buck Showalter,Orioles", 
                                     subtitle = "Buck Showalter, Baltimore Orioles")

ggplotly(showalter_plt, tooltip = c("text", "first_half_K_BB_perc_plus", "second_half_mean_gmLI"))

cooper_plt <-
  li_grouped %>%
  filter(Manager == "C.Cooper") %>%
  ggplot(aes(first_half_K_BB_perc_plus, second_half_mean_gmLI, text = pitcher_subbed_in)) + geom_point() +
  labs(x="Reliever First-Half K-BB%+", y="Reliever Second Half Mean gmLI") +
  facet_wrap(vars(Season)) + ggtitle("Decision-Making in the Second Half, Cecil Cooper", 
                                     subtitle = "Buck Showalter, Baltimore Orioles")

ggplotly(cooper_plt, tooltip = c("text", "first_half_K_BB_perc_plus", "second_half_mean_gmLI"))


roberts_plt <-
  li_grouped %>%
  filter(Manager == "D.Roberts") %>%
  ggplot(aes(first_half_K_BB_perc_plus, second_half_mean_gmLI, text = pitcher_subbed_in)) + geom_point() +
  labs(x="Reliever First-Half K-BB%+", y="Reliever Second Half Mean gmLI") +
  facet_wrap(vars(Season)) + ggtitle("Decision-Making in the Second Half, Dave Roberts", 
                                     subtitle = "Buck Showalter, Baltimore Orioles")

ggplotly(roberts_plt, tooltip = c("text", "first_half_K_BB_perc_plus", "second_half_mean_gmLI"))

```


```{r, fig.height=3}
reliever_use_plt <-
  ggplot(li_grouped, aes(first_half_K_BB_perc_plus, second_half_mean_gmLI, text = label)) +
  geom_point()

ggplotly(reliever_use_plt, tooltip = c("text", "first_half_K_BB_perc_plus", "second_half_mean_gmLI"))


top_10_names <-
  correlations %>%
  head(10)

bottom_10_names <-
  correlations %>%
  tail(10)

# graph plots of top 10 teams
top_10_plt <-
  li_grouped %>%
  unite(season_team, Season, pitching_team, sep = " ") %>%
  filter(season_team %in% top_10_names$names) %>%
  ggplot(aes(first_half_K_BB_perc_plus, second_half_mean_gmLI, text = pitcher_subbed_in)) +
  geom_point() +
  facet_wrap(vars(season_team), scales = "free_x") +
  ggtitle("Best Bullpen Usage")

bottom_10_plt <-
  li_grouped %>%
  unite(season_team, Season, pitching_team, sep = " ") %>%
  filter(season_team %in% bottom_10_names$names) %>%
  ggplot(aes(first_half_K_BB_perc_plus, second_half_mean_gmLI, text = pitcher_subbed_in)) +
  geom_point() +
  facet_wrap(vars(season_team), scales = "free_x") +
  ggtitle("Worst Bullpen Usage")

ggplotly(top_10_plt, tooltip = c("text", "first_half_K_BB_perc_plus", "second_half_mean_gmLI"))

ggplotly(bottom_10_plt, tooltip = c("text", "first_half_K_BB_perc_plus", "second_half_mean_gmLI"))

# not fair to managers really ... they can't control who's on their roster, so they can't control

# so like 2011 SFG ... Romo was much better as a pitcher than anyone else in the bullpen. yet Wilson and Affeldt got the highest leverage opportunities.
# but 2013 MIA didn't have much of a choice ...
```

