---
title: "stolen bases"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


Orioles' raw SB total would be the worst since ...

But that doesn't take league averages into account.

As we know, stolen bases have fallen out of favor. Here's a plot of stolen bases per game since 1986:


```{r}

fg_team_data <- read_csv("~/Downloads/teams.csv")

colnames(fg_team_data)[7] <- "Singles"
colnames(fg_team_data)[8] <- "Doubles"
colnames(fg_team_data)[9] <- "Triples"
colnames(fg_team_data)[1] <- "Season"




team_stolen_bases <- fg_team_data %>%
  mutate(sb_attempts_per_g = (SB + CS) / G) %>%
  inner_join(league_rates, by="Season") %>%
  mutate(sb_attempts_per_g_minus = round(sb_attempts_per_g / lg_SB_attempts_per_G * 100, 0),
         percentile = percent_rank(sb_attempts_per_g_minus) * 100) %>%
  select(Team, Season, sb_attempts_per_g, sb_attempts_per_g_minus, percentile)



```

A history of Orioles stealing bases


Some media outlets currently are focusing on the Orioles' rare attempts at stealing bases. Through X games they've attempted only Y steals, a rate of 0.0163 per game or about 1.6 per hundred games. That's about 1/4th the rate of the rest of the league, which averages six attempts per hundred games.


############
Common wisdom holds that base-stealing is down. That's true in the long run; in the past 65 years, attempted steals peaked at 10 per 100 in 1987. Rates have fallen steadily since then.

But compared to 2015, attempted steals are up a tick. Last year players bolted about five times per hundred games; this year that rate is up to about six times per game. Here's the macro-level trend:

```{r}
league_rates <- fg_team_data %>%
  filter(Season >= 1951) %>%
  group_by(Season) %>%
  summarize(lg_attempts = sum(SB) + sum(CS),
    lg_SB_attempts_per_G = lg_attempts / sum(G))

plt <- ggplot(data=league_rates, aes(x=Season, y=lg_SB_attempts_per_G)) + geom_line() + geom_point() + ggtitle("League Attempted Steals per Player-Game, 1951-2016") + labs(y="Attempts/P-G")
ggplotly(plt)
```

The Orioles refuse to follow the herd. They're running only 1.6 times per hundred games. That's about a quarter of the rest of the league. If these rates hold, it will easily be the lowest rate since the 1960 Kansas City Athletics:

[list top 10]:

```{r}

# Team Stolen Base Attempts, Percent of League Average
lowest_10_stolen_base_rates <- team_stolen_bases %>%
  arrange(sb_attempts_per_g_minus) %>%
  head(13)
lowest_10_stolen_base_rates
```

As you can see, the 2016 would "beat" the 1960 Athletics but also every other team since 1951. They'll set a new record.

The following chart shows the Orioles' own history at attempted steals, where 100 is attempts at the league-average rate:

```{r}
orioles_sb_minus <- team_stolen_bases %>%
  filter(Team == "Orioles") %>%
  mutate(above_average = ifelse(sb_attempts_per_g_minus > 100, "Yes", "No"))


orioles_sb_minus_plot <- ggplot(data=orioles_sb_minus, aes(x=Season, y=sb_attempts_per_g_minus)) + geom_line() + geom_point()  +
  ggtitle("Orioles Attempted Steals per Player-Game, 1954-2016") + labs(y="Attempts/P-G, % of League Average")

ggplotly(orioles_sb_minus_plot)
```

The first thing that jumps out is the peak in 1973. That year the Orioles attempted steals at 175% of league average. That rates at the 98th percentile since 1951.


```{r}
orioles_1973 <- read_csv("~/Downloads/FanGraphs Leaderboard (33).csv")

```


Don Baylor was a machine that year, running 41 times and getting caught only nine times. That's a 78% success rate, pretty good by any standard. Al Bumbry bolted 33 times that year, while Rich Coggins, Paul Blair, and Bobby Grich each had 26 attempts. 

1973's rate of 175% of league average was a huge jump from 1972's rate of 103% of league average. In 1972 Don Baylor attempted only 26 swipes. What caused him to nearly double his run rate? Earl Weaver, no fan of the bag swipe, managed both teams.

Perhaps the rest of the league influenced both men. Recall from the earlier graph that stolen base attempts surged between 1971 and 1976, rising from six attempts per hundred games per team to ten attempts per hundred games per team. That doesn't sound like a lot, but given the number of teams and the number of games they played, it's an increase from 2,804 attempts per year to 4,596.  

John McMurray would agree [http://sabr.org/research/examining-stolen-base-trends-decade-deadball-era-through-1970s]. He sees Baylor as part of a four-part trend in the 1970s. First, 1960's speedsters like Maury Wills and Lou Brock showed other teams the power of the swipe. (Brock himself attempted 151 steals in 1974 on his way to setting the single-season record of the time). Second, managers supported and encouraged this trend, even among hitters with a bit of power like Baylor. Third, basestealing developed as a science. Teams like the Royals began timing their runners, a practice that persists today. Finally, a league-wide decline in batting average, coupled with the introduction of large stadiums that suppressed home-run power, meant the stolen base was an attractive way to advance runners and gain an edge.

Despite their large number of attempts in '73, the Orioles have never been a running team. In 43 of the 63 years the franchise has been in Baltimore, they've attempted fewer steals than league average. Perhaps Weaver's long tenure tamped down on attempts. His emphasis on power probably shaped not only how his teams played, but who the GMs drafted. But it's curious that the talent and management philosophy required for stolen bases didn't align more often.

2007 was the last year the Orioles attempted more steals than the average team. That year a healthy Brian Roberts ran 57 times and was safe 50 of them. Remember that? Good times. Among qualified hitters only Juan Pierre, Hanley Ramirez, and Carl Crawford attempted more steals. Eric Byrnes matched Roberts' total.

Corey Patterson was also on the 2007 Orioles team and made 46 attempts, landing safely at the bag in 37 of them. Nick Markakis attempted 24 steals and was safe 18 times. Melvin Mora and Miguel Tejada hardly ran at all.

2007 concluded a seven-year run during which the Orioles' lowest attempt rate was 97% of league average. Coincidentally enough, the start of that run, the year 2000, was Mike Hargrove's first year as manager. The high attempt rate lasted through his tenure and that of Lee Mazilli and Sam Perlozzo. (Man, it feels really, really weird to type those names out.) It seems these guys pushed the team to steal more.

2008 was Dave Trembley's first full year at the helm and that's when the Orioles reverted to a below-average attempt rate. Their attempt rate dropped sharply from 142% of league average to 91%. Trembley's years at the helm bled into a few games of Juan Samuel in 2010 before Buck signed on as manager. Relative to Hargrove/Mazilli/Perlozzo, these three guys seem to want to steal less often.

In Buck's reign, 2013 was the high-water mark in terms of steal attempts. You all remember that season, right? It wasn't that long ago. The Orioles had Nate McLouth for 142 games and he bolted 37 times. Adam Jones chipped in with 17 attempts. A young(er) Manny Machado also attempted 13 steals and was thrown out in six of them. Ugh. This year, Manny's attempted only three steals and has been thrown out all three times. It's not like he can't steal bases; he made 20 of 28 attempts just last year. 

But if you're a fan of the stolen base, the Orioles aren't your team. Next year the team returns Machado, Jones, Chris Davis, J.J. Hardy, and Hyun-Soo Kim under contract. They'll probably tender contracts to Jonathan Schoop, Ryan Flaherty, and Caleb Joseph. Joey Rickard may also be back.

This is not a base-stealing roster.  






######




This refusal to run depends on the players available, sure, but contrasts with the rest of the league. Although the common wisdom is that players don't run as much as they used to, when compared to 2015 they're running *more* often. In 2015 players teams averaged ran about five times per hundred games


broadcasters are focused on the team's raw stolen base totals, you and I know that comparing raw totals

No other team has approached this low of a rate since the 1972 Tigers.

-Highest rate was 1973 when O's stole bases at 195% of league average
-Followed by '74: 161% clip


- LG SB/G rate is actually up this year from last year
- But the O's rate is down to 24% of league average
- Previous low was all the way back in 1984, when the team rate was 42% of league average




