

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rvest)
library(stringr)

fgt = theme(
  panel.background = element_rect(color = "lightgrey", fill = "white"),
  axis.title = element_text(family = "Lato"),
  axis.text = element_text(family = "Lato"),
  legend.title = element_text(family = "Lato"),
  legend.text = element_text(family = "Lato"),
  legend.background = element_rect(fill = "white"),
  legend.key = element_rect(fill = "white"),
  plot.title = element_text(family = "Lato"),
  panel.grid.major = element_line(size = .1, color = "lightgrey"),
  panel.grid.minor = element_line(size = 0),
  strip.background = element_rect(fill = "#50ae26"),
  strip.text = element_text(color="white", family="Lato")
  )

fg_db <- dplyr::src_mysql(dbname = Sys.getenv("FG_DB"), 
                            host = Sys.getenv("FG_HOST"), 
                            user = Sys.getenv("FG_USER"),
                            password = Sys.getenv("FG_PW"))
```

### Get data for starting pitchers

Starting pitchers are defined as those who started at least 25% of games they appeared in that year.

```{r}
player_lookups <- fg_db %>%
  tbl("player_info") %>%
  select(PlayerId, FirstName, LastName) %>%
  collect(n=Inf)
  
team_lookups <- fg_db %>%
  tbl("season_team") %>%
  filter(Season >= 2008) %>%
  select(TeamId, AbbName, Season) %>%
  collect(n=Inf)


pitcher_data <- fg_db %>%
  tbl("stats_pitching_master_pfx") %>%
  filter(Type >= 0,
         Type < 900,
         Season >= 2008,
         G / GS >= .25) %>%
  select(PlayerId, TeamId, Season, ValueW, `RA9-Wins`, contains("pfxvF")) %>%
  collect(n=Inf)
```  

### Combine datasets and compute team-season info

```{r}
pitcher_tbl <- pitcher_data %>%
  inner_join(player_lookups, by="PlayerId") %>%
  mutate(Name = sprintf("%s %s", FirstName, LastName)) %>%
  inner_join(team_lookups, by=c("TeamId", "Season")) %>%
  select(Name, Season, Team = AbbName, ValueW, `RA9-Wins`, contains("pfxvF")) %>%
  mutate(Name = ifelse(Name == "Lance McCullers" & Season == 2016, "REMOVEME", Name)) %>%
  filter(Name != "REMOVEME") #rosenthal was complaining b/c mccullers wasn't in the rotation yet (he was injured) 

team_tbl <- pitcher_tbl %>%
  group_by(Team, Season) %>%
  summarize(sd_fb_vel = sd(pfxvFA, na.rm = TRUE),
            mad_fb_vel = mad(pfxvFA, na.rm = TRUE),
            iqr_fb_vel = IQR(pfxvFA, na.rm = TRUE),
            mean_war = mean(ValueW),
            mean_ra9 = mean(`RA9-Wins`)) %>%
  ungroup() %>%
  mutate(mad_rank = percent_rank(desc(mad_fb_vel)) * 100)

## maybe examine relationship between sd_fb_vel and k%, bb%, hr.9, hr/fb ... 

## definitely do bayesian stuff instead of just linear relationship.
```

### Plot relationship between a rotation's fastball velocity variance and its average WAR





-- Astros' Woes Don't Include Similar Pitchers
-- No Variety in Fastaball Velocity? No problem.
-- 

... some kind of intro/tone-setting.

Last May, after the Astros had completed a miserable 7-17 April, Ken Rosenthal noted the miserable state of the team's starting rotation [http://www.foxsports.com/mlb/story/facts-and-fiction-from-the-first-month-of-the-mlb-season-050216]. He also hypothesized a cause.  I'll let him speak:

"The Astros’ starters ended April ranked last in the AL with a 5.10 ERA — and last in the majors in average fastball velocity at 89.3 mph, according to Fangraphs.

Velocity isn’t everything — the Astros’ starters were 25th last season, yet finished second in the AL in ERA. And the Rangers’ current starters are 29th in average fastball velocity but first in the AL in ERA.

Still, the Astros badly miss hard-throwing right-hander Lance McCullers, who is working his way back from a sore right shoulder.

Too many of their starters are similar.

One of them might turn it around, sure. But both? Seems doubtful."

The second-to-last line caught my eye: "Too many of their starters are similar."

Taken by itself, this is a vague statement. But note the context: Rosenthal spends the first three paragraphs talking about fastball velocity. Therefore I took his implication to be "Too many of their starters throw fastballs at a similar velocity."

I saw this and I wondered: if a rotation has guys who throw similar heat, will that rotation perform worse than a rotation with a more diverse set of arms? The logic certainly _feels_ right. I suspect the thought process is: hitting is timing. If, in a three-game series, batters see three starters who each average 93.5 MPH, hitters should be able to time fastballs from these guys more easily than they could if they were facing an 88 MPH thrower, a 93 MPH thrower, and a 97 MPH thrower. Hitters facing this more-diverse set of pitchers should have difficulty adjusting day-to-day and should perform more poorly.

It sounds true. It feels true. But is it? I decided to investigate.

METHODOLOGY

I ended up testing the following, more precise, hypothesis: within a starting rotation, as its diversity of fastball velocities decreases, the rotation's average WAR will also decrease. Specifically:

-- In the PITCH F/X era, I searched for pitchers who started at least 25% of their games in a season.
-- For these pitchers I obtained their average four-seam fastball velocity as well as their FIP- and RA9-based WARs. A few pitchers were excluded because they didn't record a four-seam fastball velocity in that season.
-- I computed the median average distance of each team-season's fastball velocities.
-- I studied how these MADs related to the WAR of the average pitcher in that rotation. I didn't use total WAR because each rotation has a different number of pitchers.

Examining the outliers gives you an idea of what kind of data I'm working with. By this method the 2011 Phillies had the most similar rotation in the dataset (MAD of 0.297 MPH); the 2011 Dodgers, the least similar (MAD of 4.862 MPH). Houston's 2016 rotation, not including Lance McCullers (whose absence Rosenthal was bemoaning), ranked very low on the similarity scale. Its MAD of 2.48 MPH, while maybe appearing similar, was very diverse. It ranked at only the 26th percentile of the 270 starting rotations in this dataset.

The following boxplots show these rotations:

```{r}

boxplot_tbl <- pitcher_tbl %>%
  filter((Team == "HOU" & Season == 2016) | (Team == "LAD" & Season == "2011") | (Team == "PHI" & Season == 2011),
         Name != "Lance McCullers") %>%
  mutate(group = sprintf("%s %s", Team, Season))

ggplot(boxplot_tbl, aes(group, pfxvFA)) + geom_boxplot() + coord_flip() + geom_jitter() + fgt +
  labs(x="", y="Pitcher's Avg. Four-Seam Fastball Velocity (MPH)") + ggtitle("Most and Least Similar Rotations by Four-Seam Fastball Velocity")

lad_2011 <- pitcher_tbl %>%
  filter(Team == "LAD",
         Season == "2011") %>%
  select(Name, pfxvFA) %>%
  arrange(desc(pfxvFA))

lad_2011

phi_2011 <- pitcher_tbl %>%
  filter(Team == "PHI",
         Season == "2011") %>%
  select(Name, pfxvFA) %>%
  arrange(desc(pfxvFA))

phi_2011



```

The Phillies that year ranged from Joe Blanton's 89.3 MPH heater to Kyle Kendrick's 93 MPH fastball. Not a huge difference there. The Dodgers, on the other hand, featured Ted Lilly, Jon Ely, Dana Eveland, and Jon Garland all throwing 88 MPH or slower -- while also rostering Rubby de la Rosa and Nathan Eovaldi who threw 94 - 95 MPH. Without McCullers, Houston in 2016 ranged from Doug Fister, tossing at 87 MPH, to Christopher Devenski and his reasonably fast 92.3 MPH heater. 

Now that you have an idea of what the data contains, we can move on to the results.

ARE ROTATIONS WITH SIMILAR FASTBALLS WORSE THAN ROTATIONS WITH DISSIMILAR FASTBALLS?

The following graph is a linear regression of a rotation's fastball diversity and its FIP-based WAR:

```{r}
ggplot(team_tbl, aes(mad_fb_vel, mean_war)) + geom_point() + geom_smooth(method="lm")

cor(team_tbl$mad_fb_vel, team_tbl$mean_war)

#r = -.0009 indicating no relationship between similarity and WAR

```

Both the scatterplot and linear regression tell us no relationship is present. That's about a straight a horizontal line as you can get; the correlation of -.001 provides more evidence. Wait a second. That's a negative sign! So Rosenthal is wrong twice. Remember I interpreted his hypothesis as "within a starting rotation, as its diversity of fastball velocities decreases, the rotation's average WAR will also decrease." First, the scatterplot and its close-to-zero correlation coefficient shows only the faintest traces of a relationship between fastball velocity diversity and the performance of other starters in the rotation. Second, this tiny relationship is the inverse of Rosenthal's hypothesis. The data show that as diversity of fastball velocities decreases (moves towards the left side of the x-axis above), WAR _increases_. If Rosenthal's hypothesis were true, the correlation coefficient would be positive.

The same (lack of) relationship holds true for RA9-based WAR:

```{r}
ggplot(team_tbl, aes(mad_fb_vel, mean_ra9)) + geom_point() + geom_smooth(method="lm")
cor(team_tbl$mad_fb_vel, team_tbl$mean_ra9)

#r = -.04, another awful one.
```

Here the line is a bit less horizontal, but a correlation coefficient of -.04 isn't going to get front offices excited.

Finally, just in case linear regression was playing tricks on me, I turned Bayesian inference and asked: given a rotation with very similar fastball velocities (MAD less than 1 MPH), what's the probability the average starter in that rotation exceeds 1.13 (the median) RA9-WAR?

I calculated this probability for all similarity buckets in the dataset:

```{r}

#| MAD FB Vel. Range | Probability of RA9-WAR >= 1.15 


bayes_results <- frame_data(
  ~`MAD Fastball Velocity Range of Rotation`, ~`Probability Avg. Starter in Rotation Exceeds 1.13 RA9-WAR`,
  "1 MPH or less", .75,
  "Between 1.1 and 2 MPH", .475,
  "Between 2.1 and 3 MPH", .460,
  "Between 3.1 and 4 MPH", .541,
  "Above 4 MPH", 1
)

bayes_results$`MAD Fastball Velocity Range of Rotation` <- factor(bayes_results$`MAD Fastball Velocity Range of Rotation`, levels = c("1 MPH or less", "Between 1.1 and 2 MPH", "Between 2.1 and 3 MPH", "Between 3.1 and 4 MPH", "Above 4 MPH"))

  


ggplot(bayes_results, aes(`MAD Fastball Velocity Range of Rotation`, `Probability Avg. Starter in Rotation Exceeds 1.13 RA9-WAR`)) + geom_bar(stat="identity") + fgt + ggtitle("A Diversity of Fastball Velocities Doesn't Help Rotations") + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The critique of Rosenthal's hypothesis stands: the strongest bucket in the set is the "less than 1 MPH bucket", meaning a rotation whose members have similar fastball velocities. While it's true the "Above 4 MPH" bucket features a 100% chance of having a good starter, keep in mind there's only one data point in the set. That's not a good sample from which to draw conclusions.

CONCLUSION






rotations whose starters have less-diverse average fastball velocities will have a lower average WAR than a rotation whose starters have more diverse 

Defining average four-seam fastball velocity is easy. Not all pitchers have four

As rotation fastball velocity similarity increases, the rotation's overall WAR decreases. put another way, as rotation similarity decreases (becomes more spread out), WAR increases. 

the data show the hypothesis is false: similarity in a rotation is, if anything, good. Rotations with lower values of MAD (higher similarity) have higher WARs than rotations with higher values of MAD (lower similarity). This relationship holds true for both FIP-base and RA9-based WAR.

more importantly, the data also show that the relationship between a rotation's fastball velocities and its performance is close to non-existent. if you're going to criticize a rotation for doing poorly, looking at how similar its pitchers' fastball velocities are isn't going to get you anywhere.


### Compute average rotation RA9-WAR for rotations in this time period, to see how FB velocity spread relates to it.

```{r}

mean_rotation_ra9 <- mean(team_tbl$mean_ra9)

#1.15
```

### Calculate P(Rotation's RA9-WAR >= 1.15 | MAD FB VEL <= 1)

```{r}



#P(B|E) = P(B) * P(E|B) / P(E)

#270 rotations in dataset

#P(B) = probability of rotation's mean ra9 equal to or exceeding 1.15
p_b <- team_tbl %>%
  filter(mean_ra9 >= 1.138461)

#135 / 270 = .5

#P(E|B) = probability of a rotation's MAD being <= 1, given its rotation's WAR exceeded 1.15

p_e_b <- p_b %>%
  filter(mad_fb_vel <= 1)

#16 / 133 = .12

#P(E) = MAD FB VEL between 0-1

p_e <- team_tbl %>%
  filter(mad_fb_vel <= 1)

#22 / 270 = .08

prob_above_avg_ra9_given_mad_fb_vel_lt_one <- (.5 * .12) / .08
prob_above_avg_ra9_given_mad_fb_vel_lt_one

#.738

```

### Calculate P(Rotation's RA9-WAR >= 1.15 | MAD FB VEL 1 > x >= 2)
```{r}
p_e_b_2 <- p_b %>%
  filter(mad_fb_vel > 1,
         mad_fb_vel <= 2)

#52 / 133 = .39

p_e_2 <- team_tbl %>%
  filter(mad_fb_vel > 1,
         mad_fb_vel <= 2)

#112 / 270 = .41

prob_above_avg_ra9_given_mad_fb_vel_bw_one_two <- (.5 * .39) / .41
prob_above_avg_ra9_given_mad_fb_vel_bw_one_two

#.468
```

### Calculate P(Rotation's RA9-WAR >= 1.15 | MAD FB VEL 2 > x >= 3)
```{r}
p_e_b_3 <- p_b %>%
  filter(mad_fb_vel > 2,
         mad_fb_vel <= 3)

#47 / 133 = .35

p_e_3 <- team_tbl %>%
  filter(mad_fb_vel > 2,
         mad_fb_vel <= 3)

#103 / 270 = .41

prob_above_avg_ra9_given_mad_fb_vel_bw_two_three <- (.5 * .35) / .38
prob_above_avg_ra9_given_mad_fb_vel_bw_two_three

#.453
```

### Calculate P(Rotation's RA9-WAR = 1 | MAD FB VEL 3 > x >= 4)


```{r}
p_e_b_4 <- p_b %>%
  filter(mad_fb_vel > 3,
         mad_fb_vel <= 4)

# 17 / 133 = .13

p_e_4 <- team_tbl %>%
  filter(mad_fb_vel > 3,
         mad_fb_vel <= 4)

#32 / 270 = .12

prob_above_avg_ra9_given_mad_fb_vel_bw_three_four <- (.5 * .13) / .12
prob_above_avg_ra9_given_mad_fb_vel_bw_three_four

#.533
```

```{r}
p_e_b_5 <- p_b %>%
  filter(mad_fb_vel > 4)

# 1 / 133 = .007

p_e_5 <- team_tbl %>%
  filter(mad_fb_vel > 4)

#1 / 270 = .003

prob_above_avg_ra9_given_mad_fb_vel_bw_three_five <- (.5 * .007) / .003
prob_above_avg_ra9_given_mad_fb_vel_bw_three_five

```


### Create tibble of probabilites

Bayesian analysis confirms what linear regression shows us: in general, rotations featuring pitchers with with similar fastball velocities do better than more diverse rotations. If a rotation's MAD fastball velocity is less than 1 MPH, the probability this rotation will average 1.13 or more RA9-WAR is 75%. As the rotation's MAD fastball velocity increases, the probability of exceeding this threshold generally decreases, hovering between 45 - 50% for the remaining buckets. The following graph shows this relationship:



FWIW, I used average RA9-WAR of a rotation instead of sum, because teams use different amounts of pitchers.

Regardless Rosenthal's hypothesis seems to be false. 

### Compute and display outlier datasets, for giggles



