

```{r}
library(tidyverse)
library(RMySQL)
library(tidytext) # for reorder_within: https://juliasilge.com/blog/reorder-within/
library(gamlss)
library(styler)
library(ggridges)
library(glue)

fgt <- theme(
  panel.background = element_rect(color = "lightgrey", fill = "white"),
  axis.title = element_text(family = "Lato"),
  axis.text = element_text(family = "Lato"),
  legend.title = element_text(family = "Lato"),
  legend.text = element_text(family = "Lato"),
  legend.background = element_rect(fill = "white"),
  legend.key = element_rect(fill = "white"),
  plot.title = element_text(family = "Lato"),
  panel.grid.major = element_line(size = .1, color = "lightgrey"),
  panel.grid.minor = element_line(size = 0),
  # strip.background = element_rect(fill = "#50ae26"),
  strip.text = element_text(family = "Lato")
)

fg_db <- dplyr::src_mysql(
  dbname = Sys.getenv("FG_DB"),
  host = Sys.getenv("FG_HOST"),
  user = Sys.getenv("FG_USER"),
  password = Sys.getenv("FG_PW")
)

TWILIO_SID <- Sys.getenv("TWILIO_SID")
TWILIO_TOKEN <- Sys.getenv("TWILIO_TOKEN")
twilio_number <- Sys.getenv("twilio_number")

send_message <- function(msg) {
  tw_send_message(from = twilio_number, to = "5123505107", body = msg)
}
```


gather data from statcast

```{r}

# get MLBAM ID's and FG Player ID's, so I can get FG stats and position info using the player's MLBAM ID
id_info <- fg_db %>%
  tbl("playerid_lookup") %>%
  dplyr::select(mlbamid, playerid) %>%
  collect(n = Inf)

# get player ID's of pitchers and whether they are a starter or reliever.
pitcher_type <- fg_db %>%
  tbl("player_info") %>%
  filter(BISPosition %in% c("SP", "RP")) %>%
  dplyr::select(playerid = PlayerId, position = BISPosition) %>%
  collect(n = Inf)

data <-
  fg_db %>%
  tbl("gd_savant_new") %>%

  # try hard to get only legit data. documentation: https://baseballsavant.mlb.com/csv-docs
  filter(
    pitch_type %in% c("FA", "FF", "FT", "FC", "FS", "SI", "SF", "SL", "CH", "CB", "CU", "KC", "KN", "EP"), # get legit pitch types only
    game_type %in% c("R", "F", "D", "L", "W"), # ignore spring training, etc. games
    zone != 0 # ignore invalid strikezone data
  ) %>%
  dplyr::select(pitcher, player_name, game_year, p_throws, pitch_type, plate_x, plate_z, release_speed, release_spin_rate, balls, strikes, zone, stand, description, xwoba) %>%
  collect(n = Inf) %>%
  inner_join(id_info, by = c("pitcher" = "mlbamid")) %>%
  inner_join(pitcher_type, by = "playerid")

# get RA9 data for pitchers for correlation purposes
pitcher_ra9 <-
  fg_db %>%
  tbl("stats_pitching") %>%
  filter(Type == 0) %>%
  dplyr::select(PlayerId, Season, ERA, R, IP) %>%
  collect(n=Inf) %>%
  mutate(ra9 = R / IP * 9) %>%
  group_by(PlayerId) %>%
  arrange(Season) %>%
  mutate(next_ra9 = lead(ra9)) %>%
  filter(!is.na(next_ra9))

```

code for emilio pagan article.

```{r}

data %>%
  filter(player_name == "Emilio Pagan") %>%
  filter(pitch_type == "FF") %>%
  ggplot(aes(plate_x, plate_z)) + geom_point() + facet_wrap(vars(game_year)) + ggtitle("Horizontal Plate Location vs. Vertical Plate Location", subtitle = "Emilio Pagan, Four-seam Fastballs") + labs(x="Pitch Horizontal Location", y = "Pitch Vertical Location", caption = "Source: Baseball Savant") + ggsave("pagan-plate-location-fastball.png")

data %>%
  filter(player_name == "Emilio Pagan") %>%
  filter(pitch_type == "FF") %>%
  group_by(game_year) %>%
  summarize(mean_vertical_height = mean(plate_z))

data %>%
  filter(player_name == "Emilio Pagan") %>%
  filter(pitch_type == "FF") %>%
  ggplot(aes(x = as.numeric(release_spin_rate))) +
  geom_histogram(binwidth = 10) +
  facet_grid(vars(game_year)) +
  ggtitle("Spin Rate vs. Number of Pitches",
    subtitle = "Emilio Pagan, Four-seam Fastballs"
  ) +
  labs(x = "Spin Rate (RPM)", y = "Number of Pitches", caption = "Source: Baseball Savant") +
  ggsave("pagan-spin-rate.png")

read_csv("~/Downloads/Pagan Pitches - Sheet1.csv") %>%
  mutate(`Whiff%` = `Whiff%` / 100) %>%
  gather(stat, value, -Year, -`Pitch Type`) %>%
  ggplot(aes(as.character(Year), value, color = stat, group = stat)) + geom_point() + geom_line() + facet_wrap(vars(`Pitch Type`)) +
  labs(x = "", caption = "Source: Baseball Savant") + ggtitle("Year vs. Effectiveness", subtitle = "Emilio Pagan, All Pitch Types") +
  ggsave("pagan-pitch-effectiveness.png")
```

Elieser Hernandez

```{r}

hernandez_gameday_data <-
  fg_db %>%
  tbl("gd_savant_new") %>%
  filter(player_name == "Elieser Hernandez",
         pitch_type %in% c("FF", "SL", "CH", "SI")) %>%
  collect(n=Inf)

data %>%
  filter(player_name == "Elieser Hernandez") %>%
  filter(pitch_type == "FF") %>%
  group_by(game_year) %>%
  summarize(mean_vertical_height = mean(plate_z))

#build tibble from statcast data and plot YoY changes. source: https://baseballsavant.mlb.com/savant-player/elieser-hernandez-622694?stats=statcast-r-pitching-mlb
hernandez_pitch_data <-
crossing(Season = c("2018", "2019"),
       pitch_type = c("FF", "SL", "CH", "SI")) %>%
  mutate(
    freq = c(.15, .621, 0, .229, .114, .536, .017, .333),
    xWOBA = c(.323, .356, NA_character_, .332, .261, .341, .271, .206),
    `Whiff%` = c(34.2, 16.1, NA_character_, 27.9, 14.5, 20.5, 9.1, 37.4),
    `% Horiz. Break vs. Average (in)` = c(11, -24, NA_character_, -25, 3, -5, -6, -4),
    `% Vert. Drop vs. Average (in)` = c(2, 12, NA_character_, -6, -6, 11, -14, -26),
    `Average Velo (MPH)` = c(82.8, 90.6, NA_character_, 80.3, 82.2, 90.6, 89.3, 78.5),
    `Average Spin Rate (RPM)` = c(1642, 2203, NA_character_, 2045, 1886, 2247, 2174, 2245),
    `Spin Axis (Degrees)` = c(250.5, 202.28, NA_character_, 119.13, 239.28, 205.5, 229.72, 142.84)) %>%
  pivot_longer(xWOBA:`Spin Axis (Degrees)`, names_to = "metric", values_to = "value", values_drop_na = TRUE, values_ptypes = list(value= double())) %>%
  mutate(weighted_val = freq * value)

# outcomes
hernandez_pitch_data %>%
  filter(metric %in% c("xWOBA", "Whiff%")) %>%
  ggplot(aes(Season, value, color = pitch_type, group = pitch_type)) + facet_wrap(vars(metric), scales = "free_y") + geom_point() + geom_line() +
  labs(y="Value", caption = "Source: Statcast") +
  scale_y_continuous(labels = function(x) (sprintf("%.3f", x))) + expand_limits(y=0) +
  ggtitle("Season vs. Whiff Rate and xWOBA", subtitle = "Elieser Hernandez")


# changes to individual pitches



# focus on changes to slider
#https://www.beyondtheboxscore.com/2019/7/10/20688274/elieser-hernandez-miami-marlins-rule-5-draft-slider-movement-caleb-smith-sandy-alcantara

# https://prospects365.com/2019/09/09/elieser-hernandez-is-worthy-of-your-consideration/

# to achieve this effectiveness, hernandez drastically improved the horizontal break on his slider compared to the break of the average pitcher's slider:

hernandez_pitch_data %>%
  filter(str_detect(metric, "vs.")) %>%
  ggplot(aes(Season, value, color = pitch_type, group = pitch_type)) + facet_wrap(vars(metric), scales = "free_y") + geom_point() + geom_line() +
  labs(y="Value", caption = "Source: Statcast") +
  scale_y_continuous(labels = function(x) (sprintf("%.3f", x))) + expand_limits(y=0) +
  ggtitle("Season vs. Horizontal and Vertical Movement vs. Average Pitch", subtitle = "Elieser Hernandez")

# velo and spin
hernandez_pitch_data %>%
  filter(str_detect(metric, "Velo|Spin")) %>%
  ggplot(aes(Season, value, color = pitch_type, group = pitch_type)) + facet_wrap(vars(metric), scales = "free_y") + geom_point() + geom_line() +
  labs(y="Value", caption = "Source: Statcast (Velo and Spin Rate), Brooks Baseball (Spin Axis)") +
  scale_y_continuous(labels = function(x) (sprintf("%.0f", x))) + expand_limits(y=0) +
  ggtitle("Season vs. Spin and Velocity", subtitle = "Elieser Hernandez")

# fastball locations
hernandez_gameday_data %>%
  filter(pitch_type == "FF") %>%
  ggplot(aes(plate_x, plate_z)) + geom_point() + facet_wrap(vars(game_year, stand)) +
  ggtitle("Season vs. Pitch Location", subtitle = "Elieser Hernandez, Fastballs") +
  labs(x = "Horizontal Pitch Location (ft)", y = "Vertical Pitch Location (ft)")

# release points
hernandez_gameday_data %>%
  ggplot(aes(release_pos_x, release_pos_z, group = pitch_type, color = pitch_type)) + 
  geom_point() + facet_wrap(vars(game_year)) + expand_limits(y = 0) + xlim(-5,5) + ylim(0, 10) +
  ggtitle("Horizontal Release Point vs. Vertical Release Point", subtitle = "Elieser Hernandez, All Pitches") +
  labs(x="Horizontal Release Point (ft)", y = "Vertical Release Point (ft)")

# show release points becoming more consistent
hernandez_gameday_data %>%
  group_by(game_year) %>%
  summarize(`Horizontal Release Point` = mad(release_pos_x),
            `Vertical Release Point` = mad(release_pos_z)) %>%
  pivot_longer(-game_year, names_to = "metric", values_to = "value") %>%
  mutate(value_in = value * 12) %>%
  ggplot(aes(as.character(game_year), value_in, group = metric, color = metric)) + geom_point() + geom_line() +
  labs(x="Season", y = "Median Absolute Deviation (in)", caption = "Source: Statcast") +
  ggtitle("Season vs. Median Absolute Deviation of Release Points", subtitle = "Elieser Hernandez")


hernandez_gameday_data %>%
  filter(pitch_type == "FF") %>%
  group_by(game_year, description) %>%
  count() %>%
  group_by(game_year) %>%
  mutate(percentage = n / sum(n) * 100) %>%
#  filter(description %in% c("called_strike", "foul", "foul_bunt", "foul_tip", "swinging_strike")) %>%
  ggplot(aes(as.character(game_year), percentage, group = description, color = description)) + geom_point() + geom_line() +
  labs(x="Season", y="Percentage of Outcomes", caption = "Source: Statcast") +
  ggtitle("Season vs. Percentage of Outcomes", subtitle = "Elieser Hernandez, Fastballs")

```



```{r}
data %>%
  filter(pitch_type == "FF") %>%
  mutate(bauer_units = as.numeric(release_spin_rate) / release_speed) %>%
  group_by(game_year) %>%
  mutate(lg_bauer_units = mean(bauer_units, na.rm = TRUE)) %>%
  group_by(game_year, pitcher, player_name) %>%
  summarize(bauer_units_plus = mean(bauer_units, na.rm = TRUE) / mean(lg_bauer_units) * 100) %>%
  group_by(game_year) %>%
  filter(game_year == 2019) %>% 
  mutate(rank = percent_rank(bauer_units_plus)) %>% 
  View()

```


group and clean data

add zone and count information for other analysis

```{r}

total_pitches_to_each_side_cutoff <- 1

# compute median xWOBA of each pitch type for both starters and relievers. we'll use this value later to determine
# whether a pitch produced a 'good' result or not
median_xwoba <-
  data %>%
  filter(
    str_detect(description, "hit_into_play"),
  ) %>%
  group_by(game_year, position, pitch_type) %>%
  summarize(median_xwoba = median(xwoba))


# add strikezone and count descriptions to data, whether the pitcher is a starter or reliever, classify pitches as good or bad,
# and compute the total number (and percentage) of good & bad pitches thrown by the pitcher

data_grouped <-
  data %>%
  filter(
    description != "foul"
  ) %>% # do not count foul balls as either good or bad
  inner_join(median_xwoba, by = c("game_year", "position", "pitch_type")) %>% # add median xWOBA for each position type
  mutate(
    zone_type_vertical = case_when(
      zone %in% c(11, 12, 1, 2, 3) ~ "High",
      zone %in% c(4, 5, 6) ~ "Middle",
      zone %in% c(7, 8, 9, 13, 14) ~ "Low"
    ),
    zone_type_horizontal = case_when(
      zone %in% c(2, 5, 8) ~ "Middle",
      stand == "R" & zone %in% c(11, 13) ~ "Inside",
      stand == "R" & zone %in% c(1, 4, 7) ~ "In",
      stand == "R" & zone %in% c(3, 6, 9) ~ "Away",
      stand == "R" & zone %in% c(12, 14) ~ "Outside",
      stand == "L" & zone %in% c(11, 13) ~ "Outside",
      stand == "L" & zone %in% c(1, 4, 7) ~ "Away",
      stand == "L" & zone %in% c(3, 6, 9) ~ "In",
      stand == "L" & zone %in% c(12, 14) ~ "Inside"
    ),
    count_type = case_when(
      balls == strikes ~ "Even",
      balls > strikes ~ "Batter Ahead",
      balls < strikes ~ "Pitcher Ahead"
    ),

    # define a good pitch as one that is either a called/swinging strike or gets hit for a sub-median xWOBA
    # remember, foul balls are not counted as either good or bad
    outcome = if_else(str_detect(description, "strike") |
      (str_detect(description, "hit_into_play") & xwoba <= median_xwoba), "good", "bad")
  ) %>%
  unite(zone_desc, zone_type_vertical, zone_type_horizontal, sep = " and ") %>%
#  unite(count, balls, strikes, sep = "-") %>%
  # group_by(player_name, game_year, p_throws, stand) %>%
  # mutate(total_pitches_to_batters_of_this_hand = n()) %>%
  # filter(total_pitches_to_batters_of_this_hand >= total_pitches_to_each_side_cutoff) %>% # try to get only full time pitchers
  group_by(pitcher, player_name, game_year, p_throws, position, count_type, zone_desc, stand, pitch_type, outcome) %>%
  summarize(
    num_pitches = n()
  ) %>%
  mutate(percentage = round(num_pitches / sum(num_pitches) * 100, 1))
#  mutate(pitch_percentage = round(num_pitches / total_pitches_to_batters_of_this_hand * 100, 1))
```



```{r}
# compute good-pitch rates for each pitcher in 2019. attach RA/9 information from FanGraphs
by_pitcher <-
  data_grouped %>%
  group_by(game_year, pitcher, player_name) %>%
  mutate(total_pitches = sum(num_pitches)) %>%
  filter(outcome == "good") %>%
  group_by(game_year, pitcher, player_name, position, total_pitches) %>%
  summarize(num_good_pitches = sum(num_pitches)) %>%
  mutate(good_pitch_percentage = round(num_good_pitches / total_pitches, 3)) %>%
  arrange(desc(good_pitch_percentage)) %>%
  inner_join(id_info, by=c("pitcher" = "mlbamid"))

# the following code uses a hierarchical beta-binomial regression model to estimate the prior expectations
# of good-pitch rates for each pitcher

# when estimating good-pitch rates, take into account the total number of pitches thrown
# good pitchers will get more chances
fit <- gamlss(cbind(num_good_pitches, total_pitches - num_good_pitches) ~ log(total_pitches) + position,
  data = by_pitcher,
  family = BB(mu.link = "identity")
)

td <- tidy(fit)

mu <- fitted(fit, parameter = "mu")
sigma <- fitted(fit, parameter = "sigma")

# compute shape parameters of posterior beta curves, given data
good_pitches_est <-
  by_pitcher %>%
  ungroup() %>%
  mutate(
    mu = mu,
    alpha0 = mu / sigma,
    beta0 = (1 - mu) / sigma,
    alpha1 = alpha0 + num_good_pitches,
    beta1 = beta0 + total_pitches - num_good_pitches,
    est_good_pitch_rate = alpha1 / (alpha1 + beta1),
    low_credible = qbeta(.025, alpha1, beta1),
    high_credible = qbeta(.975, alpha1, beta1)
  ) %>%
  left_join(pitcher_ra9, by=c("playerid" = "PlayerId", "game_year" = "Season")) %>%
  group_by(position) %>%
  arrange(desc(est_good_pitch_rate))

# show that distributions of good-pitch rates are normal
by_pitcher %>% ggplot(aes(good_pitch_percentage)) + geom_histogram() + facet_wrap(vars(position))

```

graphs and such

```{r, fig.height = 3}
# 10 players who improved the most from 2018 to 2019, grouped by position

current_year <- 2019

good_pitches_est %>%
  group_by(player_name, pitcher) %>%
  arrange(game_year, player_name, .by_group = TRUE) %>%
  mutate(good_pitch_change = est_good_pitch_rate - lag(est_good_pitch_rate)) %>%
  filter(game_year == current_year, !is.na(good_pitch_change)) %>%
  group_by(position) %>%
  top_n(10, wt = good_pitch_change) %>%
  mutate(player_name = reorder_within(player_name, good_pitch_change, position)) %>%
  ggplot(aes(reorder(player_name, good_pitch_change), good_pitch_change)) + geom_col() +
  coord_flip() + facet_wrap(vars(position), scales = "free_y") +
  scale_x_reordered() + labs(x="") +
  ggtitle(glue("Most Improved Pitchers, {current_year - 1} - {current_year}"))

# best pitchers of each type each year
good_pitches_est %>%
  group_by(game_year, position) %>%
  arrange(desc(est_good_pitch_rate, .by_group = TRUE)) %>%
  filter(row_number() <= 10) %>%
  ungroup() %>%
  unite(facet, position, game_year, sep = " ") %>%
  mutate(player_name = reorder_within(player_name, est_good_pitch_rate, facet)) %>%
  ggplot(aes(reorder(player_name, est_good_pitch_rate), est_good_pitch_rate)) + coord_flip() +
  geom_point(size = 0.7) + geom_errorbar(aes(ymin = low_credible, ymax = high_credible)) + facet_wrap(vars(facet), scales = "free_y") +
  labs(x = "", y = "Estimated Good-Pitch Rate with 95% Credible Interval") +
  ggtitle("Highest Good-Pitch Rates") +
  scale_x_reordered()

# worst pitchers of each type each year
# THIS IS NOT WORKING YET
# good_pitches_est %>%
#   group_by(game_year, position) %>%
#   arrange(est_good_pitch_rate, .by_group = TRUE) %>%
#   filter(row_number() >= 10) %>%
#   ungroup() %>%
#   unite(facet, position, game_year, sep = " ") %>%
#   mutate(player_name = reorder_within(player_name, est_good_pitch_rate, facet)) %>%
#   ggplot(aes(reorder(player_name, est_good_pitch_rate), est_good_pitch_rate)) + coord_flip() +
#   geom_point(size = 0.7) + geom_errorbar(aes(ymin = low_credible, ymax = high_credible)) + facet_wrap(vars(facet), scales = "free_y") +
#   labs(x = "", y = "Estimated Good-Pitch Rate with 95% Credible Interval") +
#   ggtitle("Lowest Good-Pitch Rates") +
#   scale_x_reordered()



```


the following code is not really appropriate anymore - it was an attempt to find pitches that needed improving, based on the pitch type, count, zone, and handedness of the batter. the data gathering and grouping above was originally intended to showcase this, but i changed it to a straight-up 'good pitch' rate instead.

to fix the graphs below I need to bring back a lot of the commented-out groupings so they line up with the graphs below.


```{r}
pitch_cutoff <- 2 # get only zones/batters who've seen at least this many pitches, from the pitcher in this count

# how to incorporate the count in it? that matters

data_grouped %>%
  filter(game_year == 2019,
         outcome == "good") %>%
  group_by(p_throws, pitch_type, count_type, zone_desc, stand) %>%
  top_n(10, wt = percentage) %>%

  

  
  
#  unite(label, player_name, pitch_type, count, zone_desc, stand, sep = " - ") %>%
 


  arrange(desc(percentage)) %>%
  head(10) %>%
  ggplot(aes(reorder(player_name, percentage), percentage)) + geom_col() + coord_flip() +
  facet_wrap(vars(pitch_type, zone_desc, stand)) +
  ggtitle("Mean xWOBA by Pitch Type, Location, and Batter Handedness",
    subtitle = glue("For pitches thrown at least {percentage_cutoff}% of the time in that location to batters of that hand
                          For pitchers with at least {total_pitches_to_each_side_cutoff * 2} pitches in 2019")
  ) +
  labs(
    x = "", y = "Percentage of Pitches that are Good",
    caption = "Good pitch: swinging strike, called strike, or batted ball with xWOBA less than the league median xWOBA"
  )

```

```{r}

data_grouped %>%
  filter(
    outcome == "good",
    num_pitches >= pitch_cutoff
  ) %>%
  unite(label, player_name, pitch_type, count, zone_desc, stand, sep = " - ") %>%
  arrange(desc(percentage)) %>%
  tail(25) %>%
  ggplot(aes(reorder(label, -percentage), percentage)) + geom_col() + coord_flip() +
  ggtitle(" by Pitch Type, Location, and Batter Handedness",
    subtitle = glue("For pitches thrown at least {percentage_cutoff}% of the time in that location to batters of that hand
                          For pitchers with at least {total_pitches_to_each_side_cutoff * 2} pitches in 2019")
  ) +
  facet_grid(vars(count)) +
  labs(x = "", y = "Percentage of Pitches that are Good")
```


