```{r setup, include=FALSE, echo=FALSE}


library(tidyverse)
library(stringr)
library(plotly)
library(RMySQL)
library(broom)

fgt = theme(
  panel.background = element_rect(color = "lightgrey", fill = "white"),
  axis.title = element_text(family = "Lato"),
  axis.text = element_text(family = "Lato"),
  legend.title = element_text(family = "Lato"),
  legend.text = element_text(family = "Lato"),
  legend.background = element_rect(fill = "white"),
  legend.key = element_rect(fill = "white"),
  plot.title = element_text(family = "Lato"),
  panel.grid.major = element_line(size = .1, color = "lightgrey"),
  panel.grid.minor = element_line(size = 0),
  strip.background = element_rect(fill = "#50ae26"),
  strip.text = element_text(color="white", family="Lato")
  )

fg_db <- dplyr::src_mysql(dbname = Sys.getenv("FG_DB"), 
                            host = Sys.getenv("FG_HOST"), 
                            user = Sys.getenv("FG_USER"),
                            password = Sys.getenv("FG_PW"))
```

### Get data

```{r}

gd_info <- fg_db %>%
  tbl("gd_roster") %>%
  select(mlbamid, playerid)

fg_info <- fg_db %>%
  tbl("player_info") %>%
  select(PlayerId, FirstName, LastName, BISPosition) %>%
  collect(n=Inf)

batting_war <- fg_db %>%
  tbl("stats_batting") %>%
  filter(Type == 0, 
         Season >= 2016) %>%
  select(playerid, Season, wrc_plus = `wRC+`) %>%
  collect(n=Inf)

savant_data_plate_coverage <- fg_db %>%
  tbl("gd_savant") %>%
  filter(game_year >= 2016) %>%
  select(batter, player_name, game_year, description, px, pz) %>%
  inner_join(gd_info, by=c("batter" = "mlbamid")) %>%
  collect(n=Inf) %>%
  filter(str_detect(description, "In play")) %>%
  mutate(px = as.numeric(px),
         pz = as.numeric(pz)) %>%
  inner_join(fg_info, by=c("playerid" = "PlayerId")) %>%
  filter(!BISPosition %in% c("", "SP", "RP")) %>%
  mutate(Name = sprintf("%s %s", FirstName, LastName),
         
         #compute inches instead of feet
         px_in = px * 12,
         pz_in = pz * 12) %>%
  inner_join(batting_war, by=c("playerid", "game_year" = "Season"))  # add wRC+

```

### Compute plate coverage info

```{r}
plate_coverage_raw <- savant_data_plate_coverage %>%
  group_by(Name, game_year, wrc_plus) %>%
  summarize(num_balls_contacted = n(),
            horiz_coverage = round(mad(px_in), 3),
            vertical_coverage = round(mad(pz_in), 3)) %>%
  mutate(total_coverage = horiz_coverage * vertical_coverage) %>%
  filter(num_balls_contacted >= 50) %>%
  group_by(game_year) %>%
  mutate(zCoverage = (total_coverage - mean(total_coverage)) / sd(total_coverage))

# can i draw total coverage for each player, madx * madz??

plate_coverage_change <- plate_coverage_raw %>%
  group_by(Name) %>%
  arrange(game_year) %>%
  mutate(total_coverage_diff = (total_coverage - lag(total_coverage)),
         wrc_diff = wrc_plus - lag(wrc_plus)) %>%
  arrange(desc(total_coverage_diff)) %>%
  filter(!is.na(total_coverage_diff))

```

NEED AN INTRO

To find whose plate coverage improved and declined the most, I looked at all fair balls in both 2016 and 2017. For each non-pitcher who hit at least 50 such balls in each season, I computed the median absolute deviation (MAD) where the pitch crossed the plate on both the x and z axes. (MLB Gameday uses z as the vertical axis because y is the measurement of a pitch's distance from home plate.) Statcast measures in feet but I report here in inches, because baseball is a game of them.

For each axis, the median absolute deviation gives a great idea of the batter's ability to hit inside/outside, or high/low, pitches while keeping them fair.

Examples: The Best and Worst of 2016

The following boxplot shows the players with the worst and best horizontal plate coverages of 2016:

```{r}
example_horiz_coverage <- savant_data_plate_coverage %>%
  filter(game_year == 2016,
         Name %in% c("Caleb Joseph", "Delino DeShields")) %>%
  select(Name, px_in)

example_vertical_coverage <- savant_data_plate_coverage %>%
  filter(game_year == 2016,
         Name %in% c("Taylor Motter", "Corey Dickerson")) %>%
  select(Name, pz_in)

```

```{r}
ggplot(example_horiz_coverage, aes(y=px_in, x=Name)) + geom_boxplot() + geom_jitter() + coord_flip()  +
  ggtitle("Worst and Best Horizontal Plate Coverage, 2016") + labs(x="", y="Horizontal Location of Pitch (inches, catcher's perspective)")
```

In 2016, Delino DeShields' bat covered the smallest amount of home plate, side-to-side, of anyone. His MAD on the x-axis (from here on out, MADx) was 0.375 feet, or 4.5 inches. This means most of the balls DeShields was able to not only hit, but also keep fair, fell within a 4.5-inch-wide horizontal section of home plate.

Conversely, Caleb Joseph had the majors' best horizontal plate coverage. At 9 inches, his MADx was twice DeShields'. Caleb was able to get to a wider variety of pitches that DeShields either couldn't hit or couldn't keep in fair territory. Caleb's excellent horizontal plate coverage makes it all the more strange that he went the entire year without an RBI [https://theringer.com/caleb-joseph-cant-buy-an-rbi-7d045924212a].)

To simplify, just look at the actual boxes in the boxplots above. You'll notice Caleb's box is much larger than DeShields', even though DeShields got to several more inside pitches than Caleb did.

Now: here's where I admit that horizontal plate coverage isn't everything. First, this measurement hides the fact that pitches not only bend and break on the way to home plate, but also have height to them. The boxplot hides both of these  Second, hitters are pitched differently. DeShields' small MADx could be because pitchers frequently threw to the same spots when facing him. Further, DeShields hit 132 balls fair in 2016; Joseph, only 106.

Notice also the small differential between the least and most horizontal plate coverage. 4.5 inches is about 1.5x the diameter of a baseball. This small differential is because home plate is a fixed distance wide. A

Taylor Motter, in his rookie year with the Rays, recorded a MADy of 0.368 feet, or 4.4 inches. If you're wondering, he's listed at 6'1". His teammate Corey Dickerson who also is 6'1" but recorded a MADy of 0.821 feet, or 9.8 inches.

```{r}
ggplot(example_vertical_coverage, aes(y=pz_in, x=Name)) + geom_boxplot() + geom_jitter()  +
  ggtitle("Worst and Best Vertical Plate Coverage, 2016") + labs(x="", y="Vertical Location of Pitch (inches)")
           
```

### Who had the best zone coverage in 2016?

Multiplaying horizontal by vertical coverage tells us who had the most overall coverage in 2016, with a minimum of 50 contacted balls. The leaders:

```{r}
top_10_zone_2016 <- plate_coverage_raw %>%
  filter(game_year == 2016) %>%
  select(Name, `Plate Coverage Distribution (sq. in)` = total_coverage) %>%
  arrange(desc(`Plate Coverage Distribution (sq. in)`))

ggplot(top_10_zone_2016 %>% head(10), aes(reorder(Name, `Plate Coverage Distribution (sq. in)`), `Plate Coverage Distribution (sq. in)`)) + geom_bar(stat="identity") + coord_flip()  + labs(x="") + ggtitle("Largest Plate Coverage, 2016")
```

Joseph and Dickerson lead the list, as does free-swinger Salvador Perez. (Former) teammates Matt Wieters and Adam Jones, another free-swinger, come in next. Except for Joseph, these guys are all full-time players. Their bats can reach a large variety of pitches in the strike zone.

Speaking of full-time players: perhaps the Orioles should consider playing Joseph more. The graph above shows that despite a miserable 6 wRC+ last year, Joseph was the best in the bigs at hitting a wide variety of pitches fair. Combined with his great defense (3.9 runs saved above average so far this year, 7th-best in MLB, per StatCorner [http://www.statcorner.com/CatcherReport.php]) he could be a useful player. The fact that his wRC+ is back to normal levels this year is only more evidence.

When thinking about plate coverage, consider that a baseball is about 2.9 inches in diameter. Joseph's plate coverage of 81 square inches means he can hit a baseball fair that's in about 28 different locations. Abreu's plate coverage of 66 inches means he can reach a baseball in about 23 different locations. Pitchers have more room to miss their spot with Abreu than they do with Joseph. These numbers aren't literal; because we're looking at median absolute deviation, there are exceptions. But the general idea holds. 

Now the 2016 laggards, those with the smallest plate coverage in 2016:

```{r}
bottom_10_zone_2016 <- plate_coverage_raw %>%
  filter(game_year == 2016) %>%
  select(Name, `Plate Coverage Distribution (sq. in)` = total_coverage) %>%
  arrange(desc(`Plate Coverage Distribution (sq. in)`))

ggplot(top_10_zone_2016 %>% tail(10), aes(reorder(Name, `Plate Coverage Distribution (sq. in)`), `Plate Coverage Distribution (sq. in)`)) + geom_bar(stat="identity") + coord_flip()  + labs(x="") + ggtitle("Smallest Plate Coverage, 2016")
```

This list contains catchers and up-the-middle players galore: Avila, Phegley, Garcia, Cervelli, La Stella, Hundley, and DeShields. Motter is a utility player but has spent most of his time at shortstop. None of these guys are particularly good hitters. Now we may have an idea as to why. Pitchers have an easier time with these guys because they can't reach as many balls. The guys in this list mostly rely on their defense to stick in the majors. Alex Avila's coverage fits about 8.6 baseballs in it.

Compariing league-worst Avila to league-best Joseph, we see the differences visually:

```{r}
# compare joseph's coverage to Avila's

joseph_avila <- savant_data_plate_coverage %>%
  filter(game_year == 2016,
         Name %in% c("Caleb Joseph", "Alex Avila")) %>%
  select(Name, px_in, pz_in)

ggplot(joseph_avila, aes(px_in, pz_in)) + geom_point() + facet_grid(~Name) + ggtitle("Locations of Pitches Hit Fair, 2016") +
  labs(x="Horizontal location (inches, catcher's perspective)", y="Vertical location (inches)")

```

Joseph clearly was able to hit, and keep fair, much wider range of pitches than Avila did.

BUT IS PLATE COVERAGE A GOOD THING?

The following graph gives us an idea:

```{r}
plate_coverage_model <- lm(plate_coverage_raw$wrc_plus ~ plate_coverage_raw$total_coverage)
summary(plate_coverage_model)

#ggplot(augment(plate_coverage_model), aes(x = .fitted, y = .resid)) + geom_point()


ggplot(plate_coverage_raw %>% filter(game_year == 2016), aes(total_coverage, wrc_plus)) + geom_point() + geom_smooth(method="lm")  + 
  ggtitle("Plate Coverage vs. wRC+, 2016") + labs(y="wRC+", x="Plate Coverage (sq. in)")


#wRC+ = 100.4 - (.0093 * Total Coverage)

```

Keeping in mind we have only eight months of data, the answer appears to be "no." The regression line slopes downwards every so slightly, indicating that not only is covering more of the plate _bad_ for players' offensive output, but also that plate coverage doesn't matter much. After all, Mike Trout and his incredibly high wRC+ (171 in 2016; 216 so far in 2017) have a plate coverage of only 53 square inches.

Although we'll have to wait for more data, this makes sense logically: plate coverage indicates simply that a player made contact with a pitch that resulted in a fair ball. Players with a wide zone may be reaching for pitches too much, resulting in weak contact and easy outs instead of letting a borderline strike or clear ball go by.

Joseph's plot above shows he made contact with pitches less than a foot off the ground. If he'd simply taken them for balls, he'd have gained an advantage over the pitcher. Perhaps instead of plate _coverage_, we should be talking about plate _discipline_.

Here's a video of him hitting this pitch: 

<div style="width:100%;height:0;padding-bottom:56%;position:relative;"><iframe src="https://giphy.com/embed/3og0ITbC1vh9rAHwWI" width="100%" height="100%" style="position:absolute" frameBorder="0" class="giphy-embed" allowFullScreen></iframe></div><p><a href="https://giphy.com/gifs/3og0ITbC1vh9rAHwWI">via GIPHY</a></p>

That's, er ... not a pitch you should be swinging at. To be fair, there were runners on the corners with no one out in the seventh inning with the Orioles down by two. He probably got too aggressive.

Let's return to the notion that Mike Trout, with his plate coverage of 53 inches, hits better than Avila or Joseph with their respective small and large coverages. Look at the graph: the points appear to form a slight curve. This would indicate a new hypothesis: you don't want to have too little plate coverage (Avila) or too much (Joseph). You want to be right in the middle (Trout).

The following graph reveals this curve:

```{r}
ggplot(plate_coverage_raw %>% filter(game_year == 2016), aes(total_coverage, wrc_plus)) + geom_point() + geom_smooth(method="loess")  + 
  ggtitle("Plate Coverage vs. wRC+, 2016") + labs(y="wRC+", x="Plate Coverage (sq. in)")
```

Our hypothesis appears validated. As plate coverage rises towards 60 square inches, wRC+ rises as well. As plate coverage increases beyond 60 square inches, wRC+ tends to drop. This curve matches our logic. You want your plate coverage to be large enough that you can hit a lot of pitches fair, but not so large you're lunging and making weak contact.

## 2017

It's too early in the season to do leadboards for plate coverage. But we can see who's doing better and worse relative to last year, keeping in mind our findings from 2016.

The following players have increased their plate coverage the most:

```{r}
# leaders
ggplot(plate_coverage_change %>% head(10), aes(reorder(Name, total_coverage_diff), total_coverage_diff)) + geom_bar(stat="identity") + coord_flip() + labs(y="Difference (sq. inches)", x="") + ggtitle("Largest Increases in Plate Coverage, 2016 – 2017")
```

Peraza's added about 32 square inches of plate coverage, enough to cover the 11 more locations of a pitched baseball. But interestingly enough, his wRC+ is down from 103 to 46. These facts provide more support for the idea that not all plate coverage is good.

Now for those who have lost the most plate coverage:

```{r}
#laggards
ggplot(plate_coverage_change %>% tail(10), aes(reorder(Name, total_coverage_diff), total_coverage_diff)) + geom_bar(stat="identity") + coord_flip()  + labs(y="Difference (sq. inches)", x="")
```

Charlie Blackmon is one of the more underrated players in baseball. He and his 122 wRC+ are having another very good year. Combined with what we know about Peraza's plate coverage and wRC+ dro, we should infer that our findings from 2016 hold true: increasing plate coverage isn't necessarily a good thing.

The following graph shows this relationship:

```{r}

ggplot(plate_coverage_change, aes(total_coverage_diff, wrc_diff)) + geom_point() + labs(y="wRC+ Difference", x="Plate Coverage Difference (sq. in)") + ggtitle("Plate Coverage Changes vs. wRC+ Changes, 2016 – 2017") + geom_smooth(method = "lm")
```

The trend is slight but clear: increasing your plate coverage may result in a small decrease in wRC+.


### Conclusion

I don't have a grand conclusion here. 


Figure out how to order these graphs by year.

```{r}

inciarte_plate_coverage <- savant_data %>%
  filter(game_year >= 2016,
         str_detect(description, "In play"),
         Name == "Ender Inciarte",
         pz > 0) %>%
  select(game_year, px, pz)

franco_plate_coverage <- savant_data %>%
  filter(game_year >= 2016,
         str_detect(description, "In play"),
         Name == "Maikel Franco",
         pz > 0) %>%
  select(game_year, px, pz)


peraza_plate_coverage <- savant_data %>%
  filter(game_year >= 2016,
         str_detect(description, "In play"),
         Name == "Jose Peraza",
         pz > 0) %>%
  select(game_year, px, pz)

shaw_plate_coverage <- savant_data %>%
  filter(game_year >= 2016,
         str_detect(description, "In play"),
         Name == "Travis Shaw",
         pz > 0) %>%
  select(game_year, px, pz)

ggplot(inciarte_plate_coverage, aes(y=px, x=as.character(game_year)))+ 
  geom_boxplot() + geom_jitter() + coord_flip()  + ggtitle("Ender Inciarte Horizontal Plate Coverage, 2016 - 2017") +
  labs(x="", y="Horizontal Location of Contact (feet)")

ggplot(franco_plate_coverage, aes(y=px, x=as.character(game_year))) + geom_boxplot() + geom_jitter() + coord_flip()  +
  ggtitle("Maikel Franco Horizontal Plate Coverage, 2016 - 2017") + labs(x="", y="Horizontal Location of Contact (feet)")

ggplot(peraza_plate_coverage, aes(y=pz, x=as.character(game_year))) + geom_boxplot() + geom_jitter()  +
  ggtitle("Jose Peraza Vertical Plate Coverage, 2016 - 2017") + labs(x="", y="Vertical Location of Contact (feet)")

ggplot(shaw_plate_coverage, aes(y=pz, x=as.character(game_year))) + geom_boxplot() + geom_jitter()  +
  ggtitle("Travis Shaw Vertical Plate Coverage, 2016 - 2017") + labs(x="", y="Vertical Location of Contact (feet)")


```