
Four Reasons to be Optimistic about Sean Manaea

Oakland A's fans are casting wary eyes at the 2017 season. The games haven't been played, but our Depth Chart projections has the team finishing [http://www.fangraphs.com/depthcharts.aspx?position=Standings] at 77-85, last place in the AL West. The team isn't bad, but the AL West became competitive seemingly overnight. The Astros have been a strong team for years now. The Angels suddenly look mildly dangerous, the Mariners supplemented their star-powered core with quality pieces, and the Rangers haven't gone away. 

If the A's do have a poor season, fans will need to focus on the bright spots. Most teams have some; as Grant Brisbee wrote at SBNation [http://www.sbnation.com/2017/2/13/14598688/atlanta-braves-brandon-phillips-trade-2017-season], even the Braves have a lot of watchable players. Hot prospects, ace pitchers, sluggers hitting dingers -- these all become reasons for fans to tune in as their team's window closes for the year.

Sean Manaea provides one such reason to watch A's games. Acquired from Kansas City in 2015 in exchange for Ben Zobrist, Manaea is a 6'5", 245-pound lefty who made his big-league debut last April and remained in the rotation the entire season. He gave up quite a few home runs, but his 14.7 K-BB% indicates he can fool batters. Give him a league-average HR/FB rate, adjust for his park and league, and his 93 xFIP- ranked alongside that of Rookie of the Year Michael Fulmer.

Manaea's not the team's ace -- that title still belongs to Sonny Gray and his projected 2.7 WAR. He isn't even their #2; Jharel Cotton's 2.4 WAR can claim that spot. But Manaea projects to earn 2.2 WAR in 162 innings, performing reliably in the middle of the rotation.

Here are four reasons why A's fans should be optimistic about Sean Manaea in 2017.

1. Good SwStr%

When batters swung at a pitch from Manaea, they missed 11.8% of the time. This rate ranks Manaea at the 91st percentile among pitchers who faced at least 100 batters last year: 

```{r}
swstr <- fg_pitchers %>%
  filter(Season == 2016) %>%
  select(Name, swstr = `SwStr%`) %>%
  arrange(desc(swstr)) %>%
  mutate(swstr = round(swstr * 100, 1),
         swstr_rank = round(percent_rank(swstr) * 100, 1)) %>%
  select(Name, swstr, swstr_rank) %>%
  filter(swstr >= 11.0,
         swstr <= 13.0) %>%
  arrange(desc(swstr_rank)) %>%
  head(25)

swstr
```

Manaea's peers are Justin Verlander, David Price, and James Paxton: good pitchers, all. Manaea got swings and misses more frequently than Madison Bumgarner and Chris Sale, among others.

What about 2017? Swinging-strike rate correlates very well year-to-year (r = 0.69), as the following graph shows:

```{r}
swstr_corr <- fg_pitchers %>%
  select(PlayerId, Name, Season, swstr = `SwStr%`) %>%
  group_by(PlayerId) %>%
  arrange(PlayerId, Season) %>%
  mutate(swstr = 100 * swstr,
    next_swstr = lead(swstr)) %>%
  filter(!is.na(swstr),
         !is.na(next_swstr)) %>%
  ungroup() %>%
  select(swstr, next_swstr)

ggplot(swstr_corr, aes(swstr, next_swstr)) + geom_point() + geom_smooth(method = "lm") + fgt

swstr_model <- lm(swstr_corr$next_swstr ~ swstr_corr$swstr)

#remember this is 2014-2016 starters who faced at least 100 batters in a season
```

2. Below-Average Walk Rate 

Manaea's walk rate isn't amazing, but at 6.2% (75th percentile), it's very good.

```{r}
walk_rate <- fg_pitchers %>%
  filter(Season == 2016) %>%
  select(Name, bb = `BB%`) %>%
  mutate(bb = round(100 * bb, 1),
         bb_rank = round(100 * percent_rank(desc(bb)), 1)) %>%
  filter(bb >= 6.1,
         bb <= 6.3) %>%
  arrange(desc(bb_rank))

walk_rate
  
```

Kevin Gausman is a good comp, as are Matt Harvey and Max Scherzer.

3. His HR/FB was 13.7%.

Wait, how is Manaea's high HR/FB rate a good thing? Because regression to the mean tells us this rate probably won't be so high next year. Regression to the mean is a basic tenet of the HR/FB stat, but just in case you need reassurance, the following graph shows year Y and year Y+1 HR/FB rates for Athletics starters since 2000 who've faced at least 100 batters in a season:    

```{r}
athletics_hr_fb <- fg_pitchers %>%
  filter(TeamId.x == 10) %>%
  select(PlayerId, Name, TeamId = TeamId.x, Season, hrfb = `HR/FB`) %>%
  group_by(PlayerId) %>%
  arrange(PlayerId, Season) %>%
  mutate(hrfb = 100 * hrfb,
    next_hrfb = lead(hrfb)) %>%
  filter(!is.na(hrfb),
         !is.na(next_hrfb))

ggplot(athletics_hr_fb, aes(hrfb, next_hrfb)) + geom_point() + geom_smooth(method = "lm") + fgt

```

This plot is random; if you squint you can make out a slight negative correlation. Either takeaway would be good news for Manaea in 2017. Incidentally, Sonny Gray better also hope regression to the mean acts on his incredibly high 2016 HR/FB rate.

4. Very good O-Swing%.

The best reason to be optimistic about Sean Manaea is this: last year he induced hitters to offer at 36.5% of his pitches out of the zone. This rate ranks in the 98th percentile among starters who faced at least 100 batters last year:

```{r}
oswing_plt_chart <- fg_pitchers %>%
  filter(Season == 2016) %>%
  select(Name, oswing = `pfxO-Swing%`) %>%
  mutate(oswing = round(100 * oswing, 1),
         oswing_rank = round(percent_rank(oswing) * 100, 1)) %>%
  arrange(desc(oswing_rank)) %>%
  head(15)

oswing_plt_chart

```

Manaea ranks just about equal with Thor here and only slightly ahead of Matt Andriese. 

(Yes, that's Manaea teammate Dillon Overton at the 94th percentile alongside Justin Verlander.)

Intuitively we know batters hit for less power when chasing pitches. But why rely in intution when we have data? Baseball Savant proves that swinging at pitches out of the zone results in a much lower ISO than swinging at piches in the zone:

```{r}
iso_by_location <- read_csv("iso on batted ball location - Sheet1.csv", 
                            col_types = list(col_character(), col_character(), col_double()))

#iso_by_location$Year = as.character(iso_by_location$Year)

ggplot(iso_by_location, aes(x=Year, y=`Avg. ISO on Swing`, color=`Pitch Location`)) + geom_line(aes(group=`Pitch Location`)) + geom_point() + scale_y_continuous(labels= function(x){
  sprintf("%.3f", x)
}) + fgt + ggtitle("Chasing Pitches is Bad for Hitters")
```

Manaea's high O-Swing% bodes well for suppressing power, especially in Oakland. This is another reason to suspect his HR/FB rate will drop next year. O-Swing% carries over well year-to-year. The relationship between a pitcher's year Y and his year Y+1 is strong:

```{r}
oswing_corr <- fg_pitchers %>%
  select(PlayerId, Name, Season, oswing = `pfxO-Swing%`) %>%
  group_by(PlayerId) %>%
  arrange(PlayerId, Season) %>%
  mutate(oswing = 100 * oswing,
    next_oswing = lead(oswing)) %>%
  filter(!is.na(oswing),
         !is.na(next_oswing)) %>%
  ungroup() %>%
  select(oswing, next_oswing)

ggplot(oswing_corr, aes(oswing, next_oswing)) + geom_point() + geom_smooth(method = "lm") + fgt
```

Manaea should continue to induce swings out of the zone, suppressing power from hitters he faces.

CONCLUSION





```{r}

# p(b|e) = p(b) * p (e|b) / p(e)
# probability (hr/fb over 11% | o-swing rate over 35%)

# b = hr/fb over 11%
# e = o-swing rate over 35%

sample <- fg_pitchers %>%
  filter(!is.na(`O-Swing%`),
         !is.na(`HR/FB`)) %>%
  select(PlayerId, Name, Season, `O-Swing%`, `HR/FB`)

#p(b)
p_b <- sample %>%
  filter(`HR/FB` >= .11)
#821 / 1769
#0.464104

#p(e|b) - given a player has a hr/fb over .11, what is the probability they had an oswing % over .35?
p_b_e <- p_b %>%
  filter(`O-Swing%` >= .350)
#24 / 821
#0.02923264

#p(e)
p_e <- sample %>%
  filter(`O-Swing%` >= .350)
#55 / 1769
#0.03109101

```










===


Getting hitters to chase pitches out of the zone correlates with success, as the following graph shows:

```{r}
oswing_plt <- fg_pitchers %>%
  filter(BISPosition == "SP",
         TBF >= 100) %>%
  select(Name, TBF, Season, oswing = `pfxO-Swing%`, ValueW) %>%
  mutate(oswing = 100 * oswing) %>%
  arrange(desc(oswing))

ggplot(oswing_plt, aes(oswing, ValueW)) + geom_point() + geom_smooth(method = "lm") + fgt + 
  labs(x="O-Swing %", y="WAR") + ggtitle("O-Swing% vs. WAR, Starting Pitchers Who Faced at least 100 Batters, 2014-2016")
```






.... where to go from here .... ?

Manaea's change-up gets the most out-of-zone swings:

http://www.fangraphs.com/leaders.aspx?pos=all&stats=pit&lg=all&qual=140&type=5&season=2016&month=0&season1=2016&ind=0&team=0&rost=0&age=0&filter=&players=0&sort=11,a

The following graph shows how this new grip, which Manaea introduced in May to combat an awful April, improved hitters' willinginess to chase his change-up out of the zone:

```{r}
manaea_ozone_change <- read_csv("manaea-changeup-oswing.csv")

manaea_ozone_change$Date <- as.Date(manaea_ozone_change$Date, format="%m-%d-%y")

ggplot(manaea_ozone_change, aes(x=Date, y=pitch_percent)) + geom_point() + geom_line() + fgt + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Back in August, Eno wrote (http://www.fangraphs.com/blogs/sean-manaea-checking-them-off/) how Manaea, a 6'5" lefty, altered how he gripped the change-up. Manaea made the shift in May  






came to a's in the trade for ben zobrist at the 2015 deadline.

- avg k%

- good bb%

- great o-swing% - 36.6%, 5th among SPs with at least 140 IP [http://www.fangraphs.com/leaders.aspx?pos=all&stats=pit&lg=all&qual=140&type=15&season=2016&month=0&season1=2016&ind=0&team=0&rost=0&age=0&filter=&players=0&sort=4,d]

- 

It's not just getting swings outside the zone. 



```{r}
fg_players <- fg_db %>%
  tbl("player_info") %>%
  collect(n=Inf)

fg_pitch <- fg_db %>%
  tbl("stats_pitching_master_pfx") %>%
  filter(Type == 0,
         Season >= 2000) %>%     #get only full-season stats for each player
  collect(n=Inf)

fg_pitchers <- inner_join(fg_players, fg_pitch, by="PlayerId") %>%
  mutate(Name = sprintf("%s %s", FirstName, LastName)) %>%
  filter(BISPosition == "SP",
         TBF >= 100)

```

```{r}
disc <- fg_pitchers %>%
  filter(BISPosition == "SP") %>%
  select(PlayerId, Name, Season, TBF, Season, oswing = `pfxO-Swing%`, zswing = `pfxZ-Swing%`, zone = `pfxZone%`,
         ocontact = `pfxO-Contact%`, ValueW, xFIP, zcontact = `pfxZ-Contact%`, swstr = `SwStr%`) %>%
  group_by(Season) %>%
  mutate(oswing_plus = 100 * (oswing / mean(oswing))) %>%
  ungroup() %>%
  filter(TBF >= 100) %>%
  mutate(ozone = 1 - zone,
         prob_ozone_and_swing = round(100 * (ozone * oswing), 1),
         prob_rank = percent_rank(prob_ozone_and_swing) * 100) %>%
  arrange(desc(prob_rank))
         
#ggplot(disc, aes(reorder(Name, prob), prob)) + geom_bar(stat="identity") + coord_flip()




```

```{r}
ggplot(disc, aes(oswing, xFIP)) + geom_point()
ggplot(disc, aes(oswstr, xFIP)) + geom_point()

oswing_model <- lm(disc$xFIP ~ disc$oswing) #r2 0.2534 
oswstr_model <- lm(disc$xFIP ~ disc$oswstr) #r2 0.3771


```
%>%
  filter(PA >= 200) %>%
  mutate(zo = zswing - oswing) %>%  #this goes across non-consecutive seasons; affects folks like Corey Patterson
  filter(is.na(zo) == FALSE) %>%  
  group_by(PlayerId) %>%
  arrange(Season) %>%
  mutate(consec = as.integer(Season) - as.integer(lag(Season))) %>%
  mutate(change = zo - lag(zo),
         wRC_change_next_year = lead(`wRC+`) - `wRC+`) %>%
  filter(!is.na(change),
         consec == 1) %>% #throw out non-consecutive seasons
  ungroup() %>%
  mutate(perc_rank = round(percent_rank(change) * 100, 2))
```
